---
title: "TX – Reconstructed Plots (Figures 2, 3, 4, 5 & Appendix A3)"
author: "Victoria"
date: "2024-10-28"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 6)

setwd('/Users/victoria/Library/CloudStorage/OneDrive-BowlingGreenStateUniversity/BGSU/PhDing/MortgageAnalysis')

packages <- c('data.table', "conflicted", "car", "forcats", "readr", "tidyverse",
              "tinytex", "datasets", "here", "data.table", "maps", "usmap",
              "stringr", "dplyr", "knitr", "parallel", "psych",
              "glmnet", "ggcorrplot", "sf", "tigris", "patchwork", "scales",
              "RColorBrewer", "viridis", "kableExtra", "cowplot")

installed_packages <- packages %in% rownames(installed.packages())
if (any(installed_packages == FALSE)) {
  install.packages(packages[!installed_packages])
}
invisible(lapply(packages, library, character.only = TRUE))
conflicts_prefer(dplyr::filter())

# ── Load and prepare data (mirrors TX_Rnd2_corrected.Rmd) ─────────────────────

data_18 <- fread('state_TX_2018.csv')
data_19 <- fread('state_TX_2019.csv')
data_20 <- fread('state_TX_2020.csv')
data_21 <- fread('state_TX_2021.csv')
data_22 <- fread('state_TX_2022.csv')
data_23 <- fread('state_TX_2023.csv')

merged_data <- rbindlist(list(data_18, data_19, data_20, data_21, data_22, data_23),
                         use.names = TRUE, fill = TRUE)

remove_cols <- function(df) {
  df %>%
    filter(loan_term == "360" & lien_status == 1 & construction_method == 1 &
           business_or_commercial_purpose == 2 & total_units == 1 &
           loan_purpose == 1 & occupancy_type == 1 &
           conforming_loan_limit == "C" & derived_race != "Free Form Text Only")
}

df <- remove_cols(merged_data)
df <- df[, -c(7,9,17,18,19,21,24:38,40,41:44,45,51:59,63:66,68:73,76,77,80:83,85:88,92)]
df[, c(1:15,21:35)] <- lapply(df[, c(1:15,21:35)], factor)

df <- df %>% filter(!(action_taken %in% c("7","8")))

df$derived_action_taken <- car::recode(df$action_taken,
  'c("1","6")="Approved"; c("2","3")="Denied"; else="Withdrawn/Incomplete"')

# Keep only completed applications (approved or denied)
df <- df %>% filter(derived_action_taken %in% c("Approved","Denied"))

df$derived_applicant_sex <- car::recode(df$applicant_sex,
  'c("1")="Male"; c("2")="Female"; c("4")="Not Applicable"; else="Other"')

df$derived_co_applicant_sex <- car::recode(df$`co-applicant_sex`,
  'c("1")="Male"; c("2")="Female"; c("4")="Not Applicable"; c("5")="No_Coapplicant"; else="Other"')

df$applicant_age <- factor(df$applicant_age,
  levels = c('<25','25-34','35-44','45-54','55-64','65-74','>74','8888'))

df <- df %>%
  mutate(race = ifelse(
    derived_ethnicity == "Joint" | derived_race == "Joint" |
      derived_sex == "Joint" | derived_co_applicant_sex %in% c(1,2,6), "Joint",
    ifelse(derived_ethnicity == "Hispanic or Latino" &
             (`applicant_race-1` %in% c(5,6,7) |
              derived_race %in% c("White","Race Not Available")), "Hispanic White",
           ifelse(`applicant_race-1` %in% c(2,27) | derived_race == "Asian", "Asian",
                  ifelse(derived_ethnicity == "Not Hispanic or Latino" &
                           (`applicant_race-1` %in% c(5,6,7) | derived_race == "White") |
                           derived_race == "White", "Non-Hispanic White",
                         ifelse(`applicant_race-1` %in% c(1,4,21:26,41:43) |
                                  derived_race %in% c("2 or more minority races",
                                    "American Indian or Alaska Native",
                                    "Native Hawaiian or Other Pacific Islander"),
                                "Other Minority Races",
                                ifelse(`applicant_race-1` == 3 |
                                         derived_race == "Black or African American",
                                       "Black or African American", "Missing")))))))

df_2 <- df %>% dplyr::select(-c(2:4,12:13,22:27,31,32,34,35))

df_3 <- df_2 %>% filter(income != 'NA') %>% filter(income != 0)
df_3 <- na.omit(df_3)
df_3[, c(12:14,21:27)] <- as.data.frame(lapply(df_3[, c(12:14,21:27)], as.numeric))

df_3$income_bins <- cut(df_3$income,
  breaks = c(0,50,100,150,200,300,400,500,600,1000,2000,Inf), include.lowest = TRUE)
df_3$log_income     <- log(df_3$income)
df_3$ltv_bins       <- cut(df_3$loan_to_value_ratio,
  breaks = c(0,60,80,90,100,150,Inf), include.lowest = TRUE)
df_3$dti_g50        <- ifelse(
  df_3$debt_to_income_ratio %in% c("<20%",'20%-<30%','30%-<36%',
    '36','37','38','39','40','41','42','43','44','45','46','47','48','49'),
  '<50%','>=50%')
df_3$property_value_thousands <- df_3$property_value / 1000
df_3$property_value_bins <- cut(df_3$property_value_thousands,
  breaks = c(0,200,400,500,1000,1500,Inf), include.lowest = TRUE)
df_3$loan_amt_thousands <- df_3$loan_amount / 1000

df_3 <- within(df_3, applicant_age       <- relevel(applicant_age,       ref='35-44'))
df_3 <- within(df_3, ltv_bins            <- relevel(ltv_bins,            ref='(90,100]'))
df_3 <- within(df_3, property_value_bins <- relevel(property_value_bins, ref='(200,400]'))
df_3 <- within(df_3, derived_sex         <- relevel(derived_sex,         ref='Joint'))
df_3$race <- factor(df_3$race)
df_3 <- within(df_3, race <- relevel(race, ref='Non-Hispanic White'))
df_3 <- df_3[complete.cases(df_3), ]
df_3$denied <- as.integer(df_3$derived_action_taken == "Denied")
```


---

# Figure 1 — County-Level Denial Rates and Annual Changes (2018–2023)

Six choropleth maps show the raw mortgage denial rate per county for each year.
Counties with fewer than 10 applications are excluded (shown in grey).
A second set of panels shows the year-over-year percentage-point change in
denial rate, matching the "annual rate changes" panel described in the manuscript.

```{r figure1_county_denial_rates, fig.height=10, fig.width=14}

options(tigris_use_cache = TRUE)
tx_counties <- counties(state = "TX", cb = TRUE, year = 2020) %>%
  mutate(county_code = paste0(STATEFP, COUNTYFP))

# ── County denial rates for every year ────────────────────────────────────────
years <- 2018:2023

county_rates <- df_3 %>%
  mutate(county_code  = as.character(county_code),
         activity_year = as.integer(as.character(activity_year))) %>%
  group_by(activity_year, county_code) %>%
  summarise(
    n_apps      = n(),
    denial_rate = mean(denied) * 100,
    .groups     = "drop"
  ) %>%
  filter(n_apps >= 10)

# ── Year-over-year change in denial rate ─────────────────────────────────────
county_changes <- county_rates %>%
  arrange(county_code, activity_year) %>%
  group_by(county_code) %>%
  mutate(delta = denial_rate - dplyr::lag(denial_rate)) %>%
  ungroup() %>%
  filter(!is.na(delta))

# ── Shared denial-rate colour scale (all years) ───────────────────────────────
rate_breaks <- c(0, 5, 10, 15, 20, 25, 30, Inf)
rate_labels <- c("0–5%","5–10%","10–15%","15–20%","20–25%","25–30%",">30%")
rate_colors <- c("#2166AC","#74ADD1","#FEE090","#FDAE61","#F46D43","#D73027","#A50026")

county_rates <- county_rates %>%
  mutate(rate_cat = cut(denial_rate, breaks = rate_breaks,
                        labels = rate_labels, include.lowest = TRUE))

# ── Panel helper: denial rate map ─────────────────────────────────────────────
make_rate_map <- function(yr) {
  map_data <- tx_counties %>%
    left_join(filter(county_rates, activity_year == yr), by = "county_code") %>%
    mutate(rate_cat = factor(rate_cat, levels = rate_labels))

  ggplot(map_data) +
    geom_sf(aes(fill = rate_cat), color = "white", size = 0.1) +
    scale_fill_manual(values  = rate_colors,
                      na.value = "#D9D9D9",
                      name    = "Denial Rate",
                      drop    = FALSE) +
    labs(title = as.character(yr)) +
    theme_void() +
    theme(plot.title      = element_text(hjust = 0.5, size = 10, face = "bold"),
          legend.position = "none")
}

# ── Panel helper: year-over-year change map ───────────────────────────────────
make_change_map <- function(yr_to) {
  map_data <- tx_counties %>%
    left_join(filter(county_changes, activity_year == yr_to), by = "county_code")

  ggplot(map_data) +
    geom_sf(aes(fill = delta), color = "white", size = 0.1) +
    scale_fill_gradient2(low      = "#2166AC",
                         mid      = "white",
                         high     = "#D73027",
                         midpoint = 0,
                         na.value = "#D9D9D9",
                         name     = "Δ pp",
                         limits   = c(-10, 10),
                         oob      = scales::squish) +
    labs(title = paste0(yr_to - 1, "→", yr_to)) +
    theme_void() +
    theme(plot.title      = element_text(hjust = 0.5, size = 9, face = "bold"),
          legend.position = "none")
}

# ── Build all panels ──────────────────────────────────────────────────────────
rate_maps   <- lapply(years,        make_rate_map)
change_maps <- lapply(2019:2023,    make_change_map)

# Shared legend for denial rate row
legend_rate <- get_legend(
  make_rate_map(2018) +
    theme(legend.position = "right",
          legend.title    = element_text(size = 8),
          legend.text     = element_text(size = 7))
)

# Shared legend for change row
legend_change <- get_legend(
  make_change_map(2019) +
    theme(legend.position = "right",
          legend.title    = element_text(size = 8),
          legend.text     = element_text(size = 7))
)

# Row 1: denial rates 2018–2023 (6 panels + legend)
row1 <- wrap_plots(rate_maps, nrow = 2) + plot_layout(guides = "collect") &
  theme(legend.position = "right",
        legend.title    = element_text(size = 8),
        legend.text     = element_text(size = 7))

# Row 2: year-over-year change (5 panels + legend)
row2 <- wrap_plots(change_maps, nrow = 2) + plot_layout(guides = "collect") &
  theme(legend.position = "right",
        legend.title    = element_text(size = 8),
        legend.text     = element_text(size = 7))
row1
# Stack both rows
row1 / row2 +
  plot_annotation(
    title   = "Figure 1: County-Level Mortgage Denial Rates and Annual Changes (Texas, 2018–2023)",
    caption = "Grey = counties with fewer than 10 applications. Change panels show percentage-point difference from prior year.",
    theme   = theme(plot.title   = element_text(hjust = 0.5, size = 12, face = "bold"),
                    plot.caption = element_text(size = 7, hjust = 0))
  )
```


---

# Figure 2 — Pearson Correlation Matrix (2018 and 2023)

The manuscript reports the Pearson correlation matrix for 2018 and 2023 across
numerical variables including log income, loan amount, property value, LTV,
minority population %, owner-occupied units, 1-to-4 family homes, and median
housing age. Darker blue = stronger positive correlation; darker red = stronger
negative association.

```{r figure2_correlation, fig.height=5, fig.width=12}

# Numerical variables used in the correlation analysis (matches manuscript Table 2)
# Correct HMDA column names confirmed from raw CSV headers


corr_vars <- c('log_income', 'loan_amt_thousands', 'property_value_thousands',
               'loan_to_value_ratio', 'tract_minority_population_percent',
               'ffiec_msa_md_median_family_income',
               'tract_to_msa_income_percentage',
               'tract_owner_occupied_units',
               'tract_one_to_four_family_homes',
               'tract_median_age_of_housing_units')

make_corr_plot <- function(yr, df) {
  dat <- df %>%
    filter(activity_year == yr) %>%
    dplyr::select(all_of(corr_vars)) %>%
    mutate(across(everything(), as.numeric)) %>%
    na.omit()

  corr_mat <- round(cor(dat, method = "pearson"), 2)

  # Shorter axis labels for readability
  labels <- c('Log Income', 'Loan Amt (k)', 'Prop Value (k)',
               'LTV', 'Minority %', 'FFIEC Income',
               'Tract/MSA Income', 'Owner Occ Units', '1-4 Family Units',
               'Median Housing Age')
  colnames(corr_mat) <- rownames(corr_mat) <- labels

  ggcorrplot(corr_mat,
             method    = "square",
             type      = "lower",
             lab       = TRUE,
             lab_size  = 2.5,
             colors    = c("#D73027", "white", "#4575B4"),
             outline.color = "grey80",
             tl.cex    = 8,
             title     = paste("Correlation Matrix —", yr)) +
    theme(plot.title = element_text(hjust = 0.5, size = 11, face = "bold"),
          legend.title = element_text(size = 8))
}

p2018 <- make_corr_plot(2018, df_3)
p2023 <- make_corr_plot(2023, df_3)

# Side-by-side layout (matches Figure 2 in manuscript)
p2018 + p2023
```


---

# Figure 3 — SIR Maps Before Risk Adjustment

County-level Standardized Incidence Ratios (SIRs) benchmarked against the
statewide denial rate (11.75% in 2018; 12.18% in 2023). The third panel shows
the change in SIR between the two years. Counties with fewer than 10 applications
are excluded (17 in 2018, 24 in 2023).

```{r figure3_sir_unadjusted, fig.height=5, fig.width=14}

# ── Texas county shapefile via tigris ─────────────────────────────────────────
options(tigris_use_cache = TRUE)
tx_counties <- counties(state = "TX", cb = TRUE, year = 2020) %>%
  mutate(county_code = paste0(STATEFP, COUNTYFP))  # 5-digit FIPS e.g. "48001"

# ── Compute county-level denial rates and raw SIRs ───────────────────────────
compute_sir_unadjusted <- function(df, yr) {
  year_data <- df %>% filter(activity_year == yr)

  # Statewide denial rate for the year
  state_rate <- mean(year_data$denied)

  county_stats <- year_data %>%
    mutate(county_code = as.character(county_code)) %>%  # factor -> character 5-digit FIPS
    group_by(county_code) %>%                  # HMDA county_code is 3-digit FIPS
    summarise(
      n_apps   = n(),
      n_denied = sum(denied),
      obs_rate = mean(denied),
      .groups  = "drop"
    ) %>%
    filter(n_apps >= 10) %>%                   # Exclude counties with <10 applications
    mutate(SIR = obs_rate / state_rate,
           year = yr)
  county_stats
}

sir_18 <- compute_sir_unadjusted(df_3, 2018)
sir_23 <- compute_sir_unadjusted(df_3, 2023)

# Join with shapefile
map_18 <- tx_counties %>%
  left_join(sir_18, by = "county_code") %>%
  mutate(SIR_label = case_when(
    is.na(SIR) ~ "Insufficient data",
    SIR < 0.75 ~ "< 0.75",
    SIR < 1.00 ~ "0.75 – 1.00",
    SIR < 1.25 ~ "1.00 – 1.25",
    SIR < 1.50 ~ "1.25 – 1.50",
    TRUE        ~ "> 1.50"))

map_23 <- tx_counties %>%
  left_join(sir_23, by = "county_code") %>%
  mutate(SIR_label = case_when(
    is.na(SIR) ~ "Insufficient data",
    SIR < 0.75 ~ "< 0.75",
    SIR < 1.00 ~ "0.75 – 1.00",
    SIR < 1.25 ~ "1.00 – 1.25",
    SIR < 1.50 ~ "1.25 – 1.50",
    TRUE        ~ "> 1.50"))

sir_levels <- c("< 0.75","0.75 – 1.00","1.00 – 1.25","1.25 – 1.50","> 1.50","Insufficient data")
sir_colors <- c("#2166AC","#92C5DE","#FDDBC7","#F4A582","#D6604D","#D9D9D9")

map_18$SIR_label <- factor(map_18$SIR_label, levels = sir_levels)
map_23$SIR_label <- factor(map_23$SIR_label, levels = sir_levels)

# Change in SIR 2018 → 2023
sir_change <- sir_18 %>%
  dplyr::select(county_code, SIR_18 = SIR) %>%
  inner_join(sir_23 %>% dplyr::select(county_code, SIR_23 = SIR), by = "county_code") %>%
  mutate(delta_SIR = SIR_23 - SIR_18)

map_change <- tx_counties %>%
  left_join(sir_change, by = "county_code")

plot_sir_map <- function(map_data, title_text) {
  ggplot(map_data) +
    geom_sf(aes(fill = SIR_label), color = "white", size = 0.15) +
    scale_fill_manual(values = sir_colors, na.value = "#D9D9D9",
                      name = "SIR", drop = FALSE) +
    labs(title = title_text) +
    theme_void() +
    theme(plot.title   = element_text(hjust = 0.5, size = 10, face = "bold"),
          legend.title = element_text(size = 8),
          legend.text  = element_text(size = 7))
}

plot_change_map <- function(map_data, title_text) {
  ggplot(map_data) +
    geom_sf(aes(fill = delta_SIR), color = "white", size = 0.15) +
    scale_fill_gradient2(low = "#2166AC", mid = "white", high = "#D6604D",
                         midpoint = 0, na.value = "#D9D9D9",
                         name = "ΔSIR") +
    labs(title = title_text) +
    theme_void() +
    theme(plot.title   = element_text(hjust = 0.5, size = 10, face = "bold"),
          legend.title = element_text(size = 8),
          legend.text  = element_text(size = 7))
}

p3a <- plot_sir_map(map_18, "SIR - 2018 (Before Adjustment)")
p3b <- plot_sir_map(map_23, "SIR - 2023 (Before Adjustment)")
p3c <- plot_change_map(map_change, "Change in SIR (2023 − 2018)")

p3c + p3a / p3b + plot_layout(ncol = 2)

p3c

```



```{r}
# counties with missing values in sir map data
length(map_18 %>% filter(is.na(SIR)) %>% pull(NAME)) #33
length(map_23 %>% filter(is.na(SIR)) %>% pull(NAME)) #34


length(map_change %>% filter(is.na(delta_SIR)) %>% pull(NAME)) #39


```

---

# Figures 4 & 5 — SIR Maps After LASSO Risk Adjustment

Using the LASSO-predicted denial probabilities as the expected rate, we compute
adjusted SIRs. Figure 4 includes all counties (showing outliers); Figure 5 removes
extreme outliers to improve interpretability.

```{r lasso_for_sir, cache=TRUE}

predictor_vars <- c('loan_amt_thousands', 'ffiec_msa_md_median_family_income',
                    'income_bins', 'race', 'derived_sex',
                    'tract_minority_population_percent',
                    'applicant_age', 'ltv_bins', 'dti_g50', 'property_value_bins')

fit_lasso <- function(yr, df) {
  yearly_data <- df %>% filter(activity_year == yr)
  x_mat <- model.matrix(
    as.formula(paste("denied ~", paste(predictor_vars, collapse = " + "))),
    data = yearly_data)[, -1]
  y_vec <- yearly_data$denied
  set.seed(123)
  cv.glmnet(x = x_mat, y = y_vec, family = "binomial", alpha = 1,
            type.measure = "mse", nfolds = 10)
}

predict_denial <- function(yr, df, cv_fit) {
  yearly_data <- df %>% filter(activity_year == yr)
  x_mat <- model.matrix(
    as.formula(paste("denied ~", paste(predictor_vars, collapse = " + "))),
    data = yearly_data)[, -1]
  pred <- as.numeric(predict(cv_fit, newx = x_mat, s = "lambda.min", type = "response"))
  yearly_data %>%
    mutate(pred_prob   = pred,
           county_code = as.character(county_code)) %>%  # factor -> character 5-digit FIPS
    dplyr::select(county_code, denied, pred_prob)
}

cat("Fitting 2018 LASSO model...\n")
fit_18 <- fit_lasso(2018, df_3)
cat("Fitting 2023 LASSO model...\n")
fit_23 <- fit_lasso(2023, df_3)

preds_18 <- predict_denial(2018, df_3, fit_18)
preds_23 <- predict_denial(2023, df_3, fit_23)
```




```{r figure4_sir_adjusted_with_outliers, fig.height=5, fig.width=10}

compute_sir_adjusted <- function(preds) {
  preds %>%
    group_by(county_code) %>%
    summarise(
      n_apps        = n(),
      obs_rate      = mean(denied),
      expected_rate = mean(pred_prob),
      .groups = "drop"
    ) %>%
    filter(n_apps >= 10) %>%
    mutate(adj_SIR = obs_rate / expected_rate)
}

adj_sir_18 <- compute_sir_adjusted(preds_18)
adj_sir_23 <- compute_sir_adjusted(preds_23)

map_adj_18 <- tx_counties %>%
  left_join(adj_sir_18, by = "county_code")
map_adj_23 <- tx_counties %>%
  left_join(adj_sir_23, by = "county_code")

# Shared fill scale (including outliers) — Figure 4
adj_range <- range(c(adj_sir_18$adj_SIR, adj_sir_23$adj_SIR), na.rm = TRUE)

plot_adj_map <- function(map_data, title_text, sir_range = NULL) {
  # Categorise adj_SIR into meaningful bins for clarity.
  # Breaks mirror the unadjusted SIR scheme so both figures are directly comparable.
  map_data <- map_data %>%
    mutate(SIR_cat = case_when(
      is.na(adj_SIR)  ~ "Insufficient data",
      adj_SIR < 0.75  ~ "< 0.75",
      adj_SIR < 1.00  ~ "0.75 – 1.00",
      adj_SIR < 1.25  ~ "1.00 – 1.25",
      adj_SIR < 1.50  ~ "1.25 – 1.50",
      adj_SIR < 2.00  ~ "1.50 – 2.00",
      TRUE            ~ "> 2.00"
    ))

  adj_levels <- c("< 0.75", "0.75 – 1.00", "1.00 – 1.25",
                  "1.25 – 1.50", "1.50 – 2.00", "> 2.00", "Insufficient data")

  # Diverging palette: blue (low denial) → white (average) → orange-red (high denial)
  # Extra bin vs Figure 3 to better distinguish counties just above 1.50
  adj_colors <- c("#2166AC", "#92C5DE", "#FDDBC7",
                  "#F4A582", "#D6604D", "#A50026", "#D9D9D9")

  map_data$SIR_cat <- factor(map_data$SIR_cat, levels = adj_levels)

  ggplot(map_data) +
    geom_sf(aes(fill = SIR_cat), color = "white", size = 0.15) +
    scale_fill_manual(values   = adj_colors,
                      na.value = "#D9D9D9",
                      name     = "Adj. SIR",
                      drop     = FALSE) +
    labs(title = title_text) +
    theme_void() +
    theme(plot.title   = element_text(hjust = 0.5, size = 10, face = "bold"),
          legend.title = element_text(size = 8),
          legend.text  = element_text(size = 7))
}

# Figure 4 — with outliers (extreme counties fall into "> 2.00" bin; no stretching)
p4a <- plot_adj_map(map_adj_18, "Adj. SIR - 2018")
p4b <- plot_adj_map(map_adj_23, "Adj. SIR - 2023")

p4a + p4b + plot_layout(ncol = 2)
```

# sir change post adjustment

```{r}
# Change in SIR 2018 → 2023
sir_change_after <- adj_sir_18 %>%
  dplyr::select(county_code, SIR_18 = adj_SIR) %>%
  inner_join(adj_sir_23 %>% dplyr::select(county_code, SIR_23 = adj_SIR), by = "county_code") %>%
  mutate(delta_SIR = SIR_23 - SIR_18)

map_change_after <- tx_counties %>%
  left_join(sir_change_after, by = "county_code")

map_change_after$adj_SIR <- map_change_after$delta_SIR

adj_range_after <- range(sir_change_after$delta_SIR, na.rm = TRUE)


p4c <- plot_change_map(map_change_after, "Change in SIR (2023 − 2018)")
p4c + p4a / p4b + plot_layout(ncol = 2)


```


# compare year to year

```{r}
sir_18_before_after <- sir_18 %>%
  dplyr::select(county_code, SIR_before = SIR) %>%
  inner_join(adj_sir_18 %>% dplyr::select(county_code, SIR_after = adj_SIR), by = "county_code") %>%
  mutate(delta_SIR = SIR_after - SIR_before)

sir_23_before_after <- sir_23 %>%
  dplyr::select(county_code, SIR_before = SIR) %>%
  inner_join(adj_sir_23 %>% dplyr::select(county_code, SIR_after = adj_SIR), by = "county_code") %>%
  mutate(delta_SIR = SIR_after - SIR_before)
```


# Visualisation A — Scatter: Unadjusted vs Adjusted SIR (Figure 3 vs Figure 4)
#
# Each dot is one county. The diagonal line = no change after adjustment.
# Above the line → adjusted SIR is HIGHER than unadjusted (structural barriers
#   not explained by credit risk). Below → credit risk accounts for the disparity.
# Outlier counties (SIR > 10 in either measure) are excluded for readability
# but annotated in the caption.

```{r sir_scatter_before_after, fig.height=5, fig.width=11}

county_names <- tx_counties %>%
  sf::st_drop_geometry() %>%
  dplyr::select(county_code, NAME)

cap_val <- 5   # cap for plot visibility; extreme outliers noted in caption

make_scatter <- function(df, yr_label) {
  d <- df %>%
    left_join(county_names, by = "county_code") %>%
    filter(!is.na(SIR_before), !is.na(SIR_after)) %>%
    mutate(
      SIR_before_capped = pmin(SIR_before, cap_val),
      SIR_after_capped  = pmin(SIR_after,  cap_val),
      direction = case_when(
        SIR_after > SIR_before * 1.05 ~ "Higher after adjustment",
        SIR_after < SIR_before * 0.95 ~ "Lower after adjustment",
        TRUE                           ~ "Approximately unchanged"
      ),
      # label only counties that moved or remain notably above 1
      label = ifelse((abs(delta_SIR) > 0.3 | SIR_after > 1.5) &
                       SIR_before <= cap_val, NAME, NA)
    )

  ggplot(d, aes(x = SIR_before_capped, y = SIR_after_capped)) +
    # reference diagonal
    geom_abline(slope = 1, intercept = 0, linetype = "dashed",
                color = "grey50", linewidth = 0.6) +
    # reference lines at SIR = 1
    geom_hline(yintercept = 1, linetype = "dotted", color = "grey70") +
    geom_vline(xintercept = 1, linetype = "dotted", color = "grey70") +
    geom_point(aes(color = direction), size = 2.2, alpha = 0.75) +
    ggrepel::geom_text_repel(aes(label = label), size = 2.4,
                              max.overlaps = 15, segment.color = "grey60") +
    scale_color_manual(
      values = c("Higher after adjustment"     = "#D6604D",
                 "Lower after adjustment"      = "#2166AC",
                 "Approximately unchanged"     = "#878787"),
      name = NULL) +
    scale_x_continuous(limits = c(0, cap_val),
                       labels = function(x) ifelse(x == cap_val,
                                                   paste0("≥", cap_val), x)) +
    scale_y_continuous(limits = c(0, cap_val),
                       labels = function(x) ifelse(x == cap_val,
                                                   paste0("≥", cap_val), x)) +
    labs(title    = paste0(yr_label, ": Unadjusted vs Risk-Adjusted SIR"),
         subtitle = "Above diagonal = disparity persists after credit risk control",
         x = "Unadjusted SIR",
         y = "Risk-Adjusted SIR") +
    theme_minimal(base_size = 10) +
    theme(legend.position  = "bottom",
          plot.title       = element_text(face = "bold", hjust = 0.5),
          plot.subtitle    = element_text(hjust = 0.5, color = "grey40", size = 8))
}

# ggrepel needed — install if missing
if (!requireNamespace("ggrepel", quietly = TRUE)) install.packages("ggrepel")
library(ggrepel)

ps_18 <- make_scatter(sir_18_before_after, "2018")
ps_23 <- make_scatter(sir_23_before_after, "2023")

ps_18 + ps_23 +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
```


# Visualisation B — Dumbbell: Top counties by adjusted SIR (2018 & 2023)
#
# Shows the TOP 20 counties ranked by their risk-ADJUSTED SIR.
# Blue dot = unadjusted (Figure 3); Red dot = adjusted (Figure 4).
# Short segments → adjustment barely changes the picture → structural barriers.
# Long segments → credit risk explained much of the raw disparity.

```{r sir_dumbbell, fig.height=7, fig.width=12}

make_dumbbell <- function(df, yr_label, n_top = 20) {
  d <- df %>%
    left_join(county_names, by = "county_code") %>%
    filter(!is.na(SIR_before), !is.na(SIR_after)) %>%
    arrange(desc(SIR_after)) %>%
    slice_head(n = n_top) %>%
    mutate(NAME = reorder(NAME, SIR_after))

  ggplot(d) +
    geom_segment(aes(x = SIR_before, xend = SIR_after,
                     y = NAME,       yend = NAME,
                     color = SIR_after > SIR_before),
                 linewidth = 1.1, show.legend = FALSE) +
    geom_point(aes(x = SIR_before, y = NAME),
               color = "#2166AC", size = 3.2) +
    geom_point(aes(x = SIR_after, y = NAME),
               color = "#D6604D", size = 3.2) +
    geom_vline(xintercept = 1, linetype = "dashed",
               color = "grey50", linewidth = 0.5) +
    scale_color_manual(values = c("TRUE" = "#D6604D", "FALSE" = "#2166AC")) +
    annotate("text", x = Inf, y = -Inf,
             label = "● Blue = unadjusted   ● Red = adjusted",
             hjust = 1.05, vjust = -0.5, size = 3, color = "grey30") +
    labs(title    = paste0(yr_label, ": Top 20 Counties - Unadjusted vs Risk-Adjusted SIR"),
         subtitle = "Dashed line = statewide SIR = 1. Short gaps = disparity not explained by credit risk.",
         x = "SIR", y = NULL) +
    theme_minimal(base_size = 10) +
    theme(plot.title    = element_text(face = "bold", hjust = 0.5),
          plot.subtitle = element_text(hjust = 0.5, color = "grey40", size = 8),
          panel.grid.major.y = element_line(color = "grey92"))
}

pd_18 <- make_dumbbell(sir_18_before_after, "2018")
pd_23 <- make_dumbbell(sir_23_before_after, "2023")

pd_18 + pd_23 + plot_layout(ncol = 2)
```


# Visualisation C — Top movers: Δ Adjusted SIR (2018→2023) + county population profiles
#
# Left panel  : lollipop of Δ adjusted SIR — which counties changed most?
# Right panel : heatmap of county-level population characteristics (minority %,
#               median income, % high DTI, median LTV) expressed as % deviation
#               from the Texas statewide average. Same county order as lollipop.
# Together they let the reader see immediately whether the counties that worsened
# share a systematic socioeconomic profile.

```{r sir_lollipop_profile, fig.height=7, fig.width=14, message=FALSE, warning=FALSE}

n_top <- 20

# ── 1. Identify top movers ────────────────────────────────────────────────────
top_movers <- sir_change_after %>%
  left_join(county_names, by = "county_code") %>%
  filter(!is.na(delta_SIR)) %>%
  mutate(abs_delta = abs(delta_SIR)) %>%
  arrange(desc(abs_delta)) %>%
  slice_head(n = n_top) %>%
  mutate(direction = ifelse(delta_SIR > 0,
                            "Worsened (2018\u21922023)",
                            "Improved (2018\u21922023)"))

top_codes <- top_movers$county_code

# ── 2. County population profiles (all years pooled for stability) ─────────────
county_profiles <- df_3 %>%
  mutate(county_code = as.character(county_code),
         income_k    = as.numeric(income),
         ltv         = as.numeric(loan_to_value_ratio),
         min_pct     = as.numeric(tract_minority_population_percent),
         high_dti    = as.integer(dti_g50 == ">=50%")) %>%
  filter(county_code %in% top_codes) %>%
  group_by(county_code) %>%
  summarise(
    `Minority\nPop. (%)` = mean(min_pct,  na.rm = TRUE),
    `Median\nIncome (k)` = median(income_k, na.rm = TRUE),
    `DTI ≥50%\n(%)`      = mean(high_dti, na.rm = TRUE) * 100,
    `Median\nLTV (%)`    = median(ltv,    na.rm = TRUE),
    .groups = "drop"
  )

# ── 3. Statewide averages for the same four measures ──────────────────────────
statewide <- df_3 %>%
  mutate(income_k = as.numeric(income),
         ltv      = as.numeric(loan_to_value_ratio),
         min_pct  = as.numeric(tract_minority_population_percent),
         high_dti = as.integer(dti_g50 == ">=50%")) %>%
  summarise(
    `Minority\nPop. (%)` = mean(min_pct,  na.rm = TRUE),
    `Median\nIncome (k)` = median(income_k, na.rm = TRUE),
    `DTI ≥50%\n(%)`      = mean(high_dti, na.rm = TRUE) * 100,
    `Median\nLTV (%)`    = median(ltv,    na.rm = TRUE)
  )

# ── 4. Compute % deviation from statewide mean for each characteristic ─────────
chars <- c("Minority\nPop. (%)", "Median\nIncome (k)", "DTI ≥50%\n(%)", "Median\nLTV (%)")

profile_long <- county_profiles %>%
  pivot_longer(all_of(chars), names_to = "characteristic", values_to = "county_val") %>%
  left_join(
    statewide %>%
      pivot_longer(everything(), names_to = "characteristic", values_to = "state_val"),
    by = "characteristic"
  ) %>%
  mutate(
    pct_dev = (county_val - state_val) / state_val * 100,
    # Fix factor order to match lollipop (ordered by delta_SIR)
    county_code = factor(county_code,
                         levels = top_movers$county_code[order(top_movers$delta_SIR)]),
    characteristic = factor(characteristic, levels = chars)
  ) %>%
  left_join(county_names, by = "county_code")

profile_long$NAME <- factor(
  profile_long$NAME,
  levels = top_movers$NAME[order(top_movers$delta_SIR)]
)

# ── 5. Left panel: lollipop ───────────────────────────────────────────────────
lollipop_data <- top_movers %>%
  mutate(NAME = factor(NAME, levels = NAME[order(delta_SIR)]))

p_lollipop <- ggplot(lollipop_data,
                     aes(x = delta_SIR, y = NAME, color = direction)) +
  geom_segment(aes(x = 0, xend = delta_SIR, y = NAME, yend = NAME),
               linewidth = 1.1) +
  geom_point(size = 3.5) +
  geom_vline(xintercept = 0, linetype = "dashed",
             color = "grey50", linewidth = 0.5) +
  scale_color_manual(
    values = c("Worsened (2018\u21922023)" = "#D6604D",
               "Improved (2018\u21922023)" = "#2166AC"),
    name = NULL) +
  labs(title    = "\u0394 Adjusted SIR",
       subtitle = "2018 \u2192 2023",
       x = "\u0394 Adj. SIR", y = NULL) +
  theme_minimal(base_size = 10) +
  theme(legend.position      = "bottom",
        plot.title           = element_text(face = "bold", hjust = 0.5),
        plot.subtitle        = element_text(hjust = 0.5, color = "grey40", size = 8),
        panel.grid.major.y   = element_line(color = "grey93"),
        axis.text.y          = element_text(size = 8))

# ── 6. Right panel: profile heatmap ──────────────────────────────────────────
p_heatmap <- ggplot(profile_long,
                    aes(x = characteristic, y = NAME, fill = pct_dev)) +
  geom_tile(color = "white", linewidth = 0.4) +
  geom_text(aes(label = round(county_val, 1)), size = 2.4, color = "grey20") +
  scale_fill_gradient2(
    low      = "#2166AC",
    mid      = "#F7F7F7",
    high     = "#D6604D",
    midpoint = 0,
    name     = "% deviation\nfrom TX avg",
    limits   = c(-100, 100),
    oob      = scales::squish
  ) +
  labs(title    = "County Population Profile",
       subtitle = "Cell value = county mean/median; colour = deviation from TX statewide avg",
       x = NULL, y = NULL) +
  theme_minimal(base_size = 10) +
  theme(plot.title         = element_text(face = "bold", hjust = 0.5),
        plot.subtitle      = element_text(hjust = 0.5, color = "grey40", size = 7),
        axis.text.y        = element_blank(),   # shared y-axis with lollipop
        axis.ticks.y       = element_blank(),
        panel.grid         = element_blank(),
        legend.title       = element_text(size = 7),
        legend.text        = element_text(size = 7))

# ── 7. Combine with patchwork ─────────────────────────────────────────────────
p_lollipop + p_heatmap +
  plot_layout(widths = c(1.4, 1), guides = "collect") +
  plot_annotation(
    title   = paste0("Top ", n_top,
                     " Counties by Change in Risk-Adjusted SIR (2018\u20232023) ",
                     "and Their Population Profiles"),
    caption = paste0(
      "Lollipop: \u0394 adjusted SIR (2023 \u2212 2018). ",
      "Heatmap: county-level mean/median for each characteristic across all years; ",
      "blue = below TX average, red = above TX average. ",
      "Characteristics derived from HMDA application-level data pooled across 2018\u20132023."),
    theme = theme(
      plot.title   = element_text(hjust = 0.5, face = "bold", size = 11),
      plot.caption = element_text(size = 7, hjust = 0)
    )
  ) &
  theme(legend.position = "bottom")
```


---

# Verification — Does minority population % predict SIR?
#
# Before writing the abstract claim, check empirically:
# 1. Is minority population % positively associated with UNADJUSTED SIR?
# 2. Does that association survive after credit risk adjustment (ADJUSTED SIR)?
#    If yes → structural discrimination story holds.
#    If the relationship collapses after adjustment → credit risk explains it.
# 3. Print Pearson r and Spearman ρ so you have numbers for the narrative.

```{r minority_sir_check, fig.height=5, fig.width=12, message=FALSE, warning=FALSE}

# ── 1. County-level average minority population % (from application-level data) ──
county_minority <- df_3 %>%
  mutate(county_code = as.character(county_code)) %>%
  group_by(county_code) %>%
  summarise(
    minority_pct = mean(as.numeric(tract_minority_population_percent), na.rm = TRUE),
    .groups = "drop"
  )

# ── 2. Build combined data frame: unadjusted & adjusted SIR + minority % ─────────
build_sir_minority <- function(sir_unadj, sir_adj, yr) {
  sir_unadj %>%
    dplyr::select(county_code, SIR_unadj = SIR) %>%
    left_join(sir_adj %>% dplyr::select(county_code, SIR_adj = adj_SIR),
              by = "county_code") %>%
    left_join(county_minority, by = "county_code") %>%
    left_join(county_names,   by = "county_code") %>%
    filter(!is.na(minority_pct), !is.na(SIR_unadj), !is.na(SIR_adj)) %>%
    # Cap extreme outliers for plotting (still counted in correlation)
    mutate(
      SIR_unadj_plot = pmin(SIR_unadj, 5),
      SIR_adj_plot   = pmin(SIR_adj,   5),
      year = as.character(yr)
    )
}

d18 <- build_sir_minority(sir_18, adj_sir_18, 2018)
d23 <- build_sir_minority(sir_23, adj_sir_23, 2023)
d_all <- bind_rows(d18, d23)

# ── 3. Correlation statistics (printed to console) ───────────────────────────────
for (yr_lbl in c("2018", "2023")) {
  d <- d_all %>% filter(year == yr_lbl)
  cat(sprintf("\n=== %s ===\n", yr_lbl))
  cat(sprintf(
    "Unadjusted SIR  vs minority %%:  Pearson r = %.3f  |  Spearman ρ = %.3f\n",
    cor(d$SIR_unadj,  d$minority_pct, use = "complete.obs", method = "pearson"),
    cor(d$SIR_unadj,  d$minority_pct, use = "complete.obs", method = "spearman")
  ))
  cat(sprintf(
    "Adjusted SIR    vs minority %%:  Pearson r = %.3f  |  Spearman ρ = %.3f\n",
    cor(d$SIR_adj,    d$minority_pct, use = "complete.obs", method = "pearson"),
    cor(d$SIR_adj,    d$minority_pct, use = "complete.obs", method = "spearman")
  ))
}

# ── 4. Scatter plots: minority % on x-axis, SIR on y-axis ────────────────────────
make_minority_scatter <- function(d, yr_lbl) {
  # Label notable counties: high SIR or high minority %
  d <- d %>%
    mutate(label = ifelse(SIR_unadj_plot >= 1.8 | SIR_adj_plot >= 1.8 |
                            minority_pct >= 70, NAME, NA))

  p_unadj <- ggplot(d, aes(x = minority_pct, y = SIR_unadj_plot)) +
    geom_hline(yintercept = 1, linetype = "dotted", color = "grey60") +
    geom_point(alpha = 0.65, color = "#4575B4", size = 2) +
    geom_smooth(method = "lm", se = TRUE, color = "#D73027",
                linewidth = 0.8, fill = "#FDAE61", alpha = 0.2) +
    ggrepel::geom_text_repel(aes(label = label), size = 2.3,
                              max.overlaps = 12, segment.color = "grey60") +
    scale_y_continuous(labels = function(x) ifelse(x == 5, "≥5", x)) +
    labs(title    = paste0(yr_lbl, ": Unadjusted SIR"),
         subtitle = paste0("r = ",
           round(cor(d$SIR_unadj, d$minority_pct,
                     use = "complete.obs", method = "pearson"), 2)),
         x = "Minority Population (%)", y = "Unadjusted SIR") +
    theme_minimal(base_size = 10) +
    theme(plot.title    = element_text(face = "bold", hjust = 0.5),
          plot.subtitle = element_text(hjust = 0.5, color = "grey40", size = 9))

  p_adj <- ggplot(d, aes(x = minority_pct, y = SIR_adj_plot)) +
    geom_hline(yintercept = 1, linetype = "dotted", color = "grey60") +
    geom_point(alpha = 0.65, color = "#D6604D", size = 2) +
    geom_smooth(method = "lm", se = TRUE, color = "#4575B4",
                linewidth = 0.8, fill = "#92C5DE", alpha = 0.2) +
    ggrepel::geom_text_repel(aes(label = label), size = 2.3,
                              max.overlaps = 12, segment.color = "grey60") +
    scale_y_continuous(labels = function(x) ifelse(x == 5, "≥5", x)) +
    labs(title    = paste0(yr_lbl, ": Risk-Adjusted SIR"),
         subtitle = paste0("r = ",
           round(cor(d$SIR_adj, d$minority_pct,
                     use = "complete.obs", method = "pearson"), 2)),
         x = "Minority Population (%)", y = "Adjusted SIR") +
    theme_minimal(base_size = 10) +
    theme(plot.title    = element_text(face = "bold", hjust = 0.5),
          plot.subtitle = element_text(hjust = 0.5, color = "grey40", size = 9))

  list(unadj = p_unadj, adj = p_adj)
}

plots_18 <- make_minority_scatter(d18, "2018")
plots_23 <- make_minority_scatter(d23, "2023")

# Layout: unadjusted 2018 | adjusted 2018 | unadjusted 2023 | adjusted 2023
(plots_18$unadj + plots_18$adj + plots_23$unadj + plots_23$adj) +
  plot_layout(ncol = 4) +
  plot_annotation(
    title   = "Minority Population Share vs SIR — Before and After Credit Risk Adjustment",
    caption = paste0("Each point = one Texas county. Red/blue line = OLS fit with 95% CI. ",
                     "Counties with SIR > 5 shown at 5 for scale. r = Pearson correlation."),
    theme   = theme(plot.title   = element_text(hjust = 0.5, face = "bold", size = 12),
                    plot.caption = element_text(size = 7, hjust = 0))
  )
```


#

# there were no outliers - after some preprocessing

```{r figure5_sir_adjusted_no_outliers, fig.height=5, fig.width=10}

# Outlier thresholds per manuscript:
# 2018: Winkler (134), Madison (46), Jim Wells (17), Bandera (16) — exclude SIR > 10
# 2023: 8 counties exceeded SIR of 10 — exclude SIR > 10
outlier_threshold <- 10

adj_sir_18_clean <- adj_sir_18 %>% filter(adj_SIR <= outlier_threshold)
adj_sir_23_clean <- adj_sir_23 %>% filter(adj_SIR <= outlier_threshold)

map_adj_18_clean <- tx_counties %>%
  left_join(adj_sir_18_clean, by = "county_code")
map_adj_23_clean <- tx_counties %>%
  left_join(adj_sir_23_clean, by = "county_code")

clean_range <- range(c(adj_sir_18_clean$adj_SIR, adj_sir_23_clean$adj_SIR), na.rm = TRUE)

# Figure 5 — without outliers; manuscript shows 2023 on left, 2018 on right
p5a <- plot_adj_map(map_adj_23_clean, "Adj. SIR - 2023 (outliers removed)")
p5b <- plot_adj_map(map_adj_18_clean, "Adj. SIR - 2018 (outliers removed)")

p5b + p5a + plot_layout(ncol = 2)
```


---

# Appendix A3 — Denial Rates by Race and Income

```{r appendixA3_denial_by_race_income, fig.height=8, fig.width=12}

# ── Panel A: Denial rate by Race and Year ────────────────────────────────────
race_denial <- df_3 %>%
  filter(race != "Missing") %>%
  group_by(activity_year, race) %>%
  summarise(denial_rate = mean(denied) * 100, .groups = "drop") %>%
  mutate(activity_year = as.integer(as.character(activity_year)))

race_order <- c("Non-Hispanic White", "Hispanic White", "Asian",
                "Black or African American", "Other Minority Races", "Joint")

race_denial$race <- factor(race_denial$race, levels = race_order)

p_race <- ggplot(race_denial, aes(x = activity_year, y = denial_rate,
                                   color = race, group = race)) +
  geom_line(size = 1) +
  geom_point(size = 2.5) +
  scale_color_brewer(palette = "Set1", name = "Race") +
  scale_x_continuous(breaks = 2018:2023) +
  labs(title = "Denial Rate by Race (2018–2023)",
       x = "Year", y = "Denial Rate (%)") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "bottom",
        plot.title = element_text(hjust = 0.5, face = "bold"))

# ── Correct income bin order (matches cut() breaks, low → high) ───────────────
income_bin_order <- c("[0,50]", "(50,100]", "(100,150]", "(150,200]",
                      "(200,300]", "(300,400]", "(400,500]", "(500,600]",
                      "(600,1e+03]", "(1e+03,2e+03]", "(2e+03,Inf]")

# ── Panel B: Denial rate by Income Bin and Year ───────────────────────────────
income_denial <- df_3 %>%
  group_by(activity_year, income_bins) %>%
  summarise(denial_rate = mean(denied) * 100, .groups = "drop") %>%
  mutate(activity_year = as.integer(as.character(activity_year))) %>%
  filter(!is.na(income_bins)) %>%
  mutate(income_bins = factor(income_bins, levels = income_bin_order))

p_income <- ggplot(income_denial, aes(x = activity_year, y = denial_rate,
                                       color = income_bins, group = income_bins)) +
  geom_line(size = 1) +
  geom_point(size = 2.5) +
  scale_color_viridis_d(name = "Income Bin (k)", option = "D") +
  scale_x_continuous(breaks = 2018:2023) +
  labs(title = "Denial Rate by Income Bin (2018–2023)",
       x = "Year", y = "Denial Rate (%)") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "bottom",
        plot.title = element_text(hjust = 0.5, face = "bold"))

# ── Panel C: Denial rate by Race × Income bin (most recent year, 2023) ────────
# Only show the five lowest income bins where most applicants are concentrated
panel_c_bins <- income_bin_order

race_income_23 <- df_3 %>%
  filter(activity_year == 2023, race != "Missing",
         income_bins %in% panel_c_bins) %>%
  group_by(race, income_bins) %>%
  summarise(denial_rate = mean(denied) * 100, n = n(), .groups = "drop") %>%
  filter(n >= 20) %>%
  mutate(
    race       = factor(race,       levels = race_order),
    income_bins = factor(income_bins, levels = panel_c_bins)   # explicit low→high order
  )

p_race_income <- ggplot(race_income_23,
                        aes(x = income_bins, y = denial_rate, fill = race)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  scale_fill_brewer(palette = "Set1", name = "Race") +
  labs(title = "Denial Rate by Race × Income Bin (2023)",
       x = "Income Bin (k)", y = "Denial Rate (%)") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "bottom",
        axis.text.x     = element_text(angle = 30, hjust = 1),
        plot.title      = element_text(hjust = 0.5, face = "bold"))

# Combine all three panels
(p_race + p_income) / p_race_income + plot_layout(heights = c(1, 1.2))
```


---

# Appendix A4 — EDA: County Characteristics by SIR Category
#
# Supports the claim: "Counties with persistently lower SIRs tend to have lower
# median property values, newer housing units, and higher proportions of minority
# residents. Minority applicants in these counties experience lower denial rates."
#
# Layout:
#   Four box plots (one per characteristic), each faceted by SIR type
#   (Unadjusted vs Risk-Adjusted) with 2018 and 2023 overlaid as fill.
#   A fifth panel shows minority applicant denial rates across the same groups.

```{r appendix_a4_sir_eda, fig.height=12, fig.width=13, message=FALSE, warning=FALSE}

# ── 1. County-level characteristics from individual applications ──────────────
county_chars_eda <- df_3 %>%
  mutate(
    county_code      = as.character(county_code),
    activity_year    = as.integer(as.character(activity_year)),
    property_val_k   = as.numeric(property_value) / 1000,
    housing_age      = as.numeric(tract_median_age_of_housing_units),
    minority_pct     = as.numeric(tract_minority_population_percent),
    is_minority      = race %in% c("Black or African American", "Hispanic White",
                                    "Asian", "Other Minority Races")
  ) %>%
  filter(activity_year %in% c(2018, 2023)) %>%
  group_by(county_code, activity_year) %>%
  summarise(
    median_prop_val    = median(property_val_k,  na.rm = TRUE),
    median_housing_age = median(housing_age,     na.rm = TRUE),
    mean_minority_pct  = mean(minority_pct,      na.rm = TRUE),
    minority_denial_rt = mean(denied[is_minority], na.rm = TRUE) * 100,
    n_minority         = sum(is_minority),
    .groups = "drop"
  ) %>%
  # exclude counties with fewer than 10 minority applicants (unstable rates)
  mutate(minority_denial_rt = ifelse(n_minority < 10, NA, minority_denial_rt))

# ── 2. Stack unadjusted and adjusted SIR for both years ──────────────────────
sir_stack <- bind_rows(
  sir_18     %>% mutate(year = 2018L, sir_type = "Unadjusted") %>%
    dplyr::select(county_code, year, sir_type, SIR),
  sir_23     %>% mutate(year = 2023L, sir_type = "Unadjusted") %>%
    dplyr::select(county_code, year, sir_type, SIR),
  adj_sir_18 %>% mutate(year = 2018L, sir_type = "Risk-Adjusted", SIR = adj_SIR) %>%
    dplyr::select(county_code, year, sir_type, SIR),
  adj_sir_23 %>% mutate(year = 2023L, sir_type = "Risk-Adjusted", SIR = adj_SIR) %>%
    dplyr::select(county_code, year, sir_type, SIR)
)

# ── 3. SIR categories (three bands centred on 1) ─────────────────────────────
sir_cat_levels <- c("Low (< 0.75)", "Medium (0.75\u20131.25)", "High (> 1.25)")

sir_stack <- sir_stack %>%
  mutate(
    SIR_cat  = case_when(
      SIR <  0.75 ~ "Low (< 0.75)",
      SIR <= 1.25 ~ "Medium (0.75\u20131.25)",
      SIR >  1.25 ~ "High (> 1.25)",
      TRUE        ~ NA_character_
    ),
    SIR_cat  = factor(SIR_cat,  levels = sir_cat_levels),
    sir_type = factor(sir_type, levels = c("Unadjusted", "Risk-Adjusted")),
    year     = as.integer(as.character(year))  # keep integer for join; convert to factor after
  )

# ── 4. Join with county characteristics, then convert year to factor for plots ─
eda_full <- sir_stack %>%
  left_join(county_chars_eda, by = c("county_code", "year" = "activity_year")) %>%
  filter(!is.na(SIR_cat)) %>%
  mutate(year = factor(year))

# ── 5. Shared theme and palette ───────────────────────────────────────────────
yr_colors  <- c("2018" = "#4575B4", "2023" = "#D73027")
box_theme  <- theme_minimal(base_size = 10) +
  theme(
    legend.position    = "bottom",
    # Solid border around each facet panel — creates a clear box per SIR type
    panel.border       = element_rect(colour = "grey30", fill = NA, linewidth = 1),
    # Coloured strip header so Unadjusted and Risk-Adjusted are immediately distinct
    strip.background   = element_rect(fill = "#D9E8F5", colour = "grey30", linewidth = 1),
    strip.text         = element_text(face = "bold", size = 9, colour = "grey10"),
    # Extra horizontal breathing room between the two facets
    panel.spacing.x    = unit(1.2, "lines"),
    axis.text.x        = element_text(size = 8),
    plot.title         = element_text(face = "bold", hjust = 0.5, size = 10),
    panel.grid.major.x = element_blank()
  )

make_box <- function(data, y_var, y_label, title_text, y_cap = NULL) {
  p <- ggplot(data, aes(x = SIR_cat, y = .data[[y_var]], fill = year)) +
    geom_boxplot(outlier.size = 0.8, outlier.alpha = 0.4,
                 position = position_dodge(0.8), width = 0.65, alpha = 0.85) +
    facet_wrap(~sir_type, ncol = 2) +
    scale_fill_manual(values = yr_colors, name = "Year") +
    labs(title = title_text, x = "SIR Category", y = y_label) +
    box_theme

  if (!is.null(y_cap))
    p <- p + coord_cartesian(ylim = c(0, y_cap))
  p
}

# ── 6. Individual panels ──────────────────────────────────────────────────────

p_propval <- make_box(eda_full, "median_prop_val",
                      "Median Property Value (k)",
                      "A. Property Value by SIR Category",
                      y_cap = 1500)

p_age     <- make_box(eda_full, "median_housing_age",
                      "Median Age of Housing Units (yrs)",
                      "B. Housing Age by SIR Category",
                      y_cap = 60)

p_minpct  <- make_box(eda_full, "mean_minority_pct",
                      "Minority Population (%)",
                      "C. Minority Population Share by SIR Category")

p_mindenial <- make_box(
  eda_full %>% filter(!is.na(minority_denial_rt)),
  "minority_denial_rt",
  "Denial Rate - Minority Applicants (%)",
  "D. Minority Applicant Denial Rate by SIR Category",
  y_cap = 60
)

# ── 7. Combine all four panels ────────────────────────────────────────────────
(p_propval + p_age) / (p_minpct + p_mindenial) +
  plot_layout(guides = "collect") +
  plot_annotation(
    title   = paste0("County Characteristics and Minority Denial Rates ",
                     "by SIR Category - 2018 and 2023"),
    # caption = paste0(
    #   "Each box = distribution across Texas counties in that SIR band. ",
    #   "Left facet = unadjusted SIR; right facet = risk-adjusted SIR. ",
    #   "Panel D excludes counties with fewer than 10 minority applicants. ",
    #   "Property values capped at $1.5M; housing age capped at 60 yrs for readability. ",
    #   "SIR categories: Low < 0.75 (below statewide average), ",
    #   "Medium 0.75\u20131.25 (near average), High > 1.25 (above average)."),
    theme = theme(
      plot.title   = element_text(hjust = 0.5, face = "bold", size = 12),
      plot.caption = element_text(size = 7, hjust = 0),
      legend.position = "bottom"
    )
  ) &
  theme(legend.position = "bottom")
```


# Appendix A4b — Persistence: Counties Consistently Low vs High SIR (Both Years)
#
# Isolates counties that are *persistently* low (SIR < 0.75 in BOTH 2018 and 2023)
# vs persistently high (SIR > 1.25 in BOTH years) and compares their profiles
# directly — the strongest evidence for the "persistently lower" claim.

```{r appendix_a4b_persistence, fig.height=6, fig.width=13, message=FALSE, warning=FALSE}

classify_persistent <- function(sir_df_18, sir_df_23, sir_col = "SIR") {
  sir_df_18 %>%
    dplyr::select(county_code, SIR_18 = all_of(sir_col)) %>%
    inner_join(
      sir_df_23 %>% dplyr::select(county_code, SIR_23 = all_of(sir_col)),
      by = "county_code"
    ) %>%
    mutate(
      persistence = case_when(
        SIR_18 < 0.75 & SIR_23 < 0.75  ~ "Persistently Low\n(both years < 0.75)",
        SIR_18 > 1.25 & SIR_23 > 1.25  ~ "Persistently High\n(both years > 1.25)",
        TRUE                             ~ "Variable / Middle"
      )
    ) %>%
    filter(persistence != "Variable / Middle")
}

persist_unadj <- classify_persistent(sir_18, sir_23, "SIR") %>%
  mutate(sir_type = "Unadjusted")

persist_adj   <- classify_persistent(
  adj_sir_18 %>% rename(SIR = adj_SIR),
  adj_sir_23 %>% rename(SIR = adj_SIR),
  "SIR"
) %>%
  mutate(sir_type = "Risk-Adjusted")

persist_all <- bind_rows(persist_unadj, persist_adj) %>%
  mutate(sir_type = factor(sir_type, levels = c("Unadjusted", "Risk-Adjusted")))

# Get 2023 county characteristics for persistent counties
persist_chars <- persist_all %>%
  left_join(county_chars_eda %>% filter(activity_year == 2023),
            by = "county_code") %>%
  filter(!is.na(median_prop_val)) %>%
  mutate(persistence = factor(persistence,
    levels = c("Persistently Low\n(both years < 0.75)",
               "Persistently High\n(both years > 1.25)")))

persist_colors <- c(
  "Persistently Low\n(both years < 0.75)"  = "#2166AC",
  "Persistently High\n(both years > 1.25)" = "#D6604D"
)

make_persist_box <- function(y_var, y_label, title_text, y_cap = NULL) {
  p <- ggplot(persist_chars,
              aes(x = persistence, y = .data[[y_var]], fill = persistence)) +
    geom_boxplot(outlier.size = 0.9, outlier.alpha = 0.5,
                 width = 0.5, alpha = 0.85, show.legend = FALSE) +
    geom_jitter(width = 0.12, size = 1, alpha = 0.4, color = "grey30") +
    facet_wrap(~sir_type, ncol = 2) +
    scale_fill_manual(values = persist_colors) +
    labs(title = title_text, x = NULL, y = y_label) +
    box_theme +
    theme(axis.text.x = element_text(size = 8, lineheight = 0.85))

  if (!is.null(y_cap))
    p <- p + coord_cartesian(ylim = c(0, y_cap))
  p
}

pp1 <- make_persist_box("median_prop_val",    "Median Property Value (k)",
                        "A. Property Value",   y_cap = 1500)
pp2 <- make_persist_box("median_housing_age",  "Median Housing Age (yrs)",
                        "B. Housing Age",       y_cap = 60)
pp3 <- make_persist_box("mean_minority_pct",   "Minority Population (%)",
                        "C. Minority Share")
pp4 <- make_persist_box("minority_denial_rt",  "Minority Denial Rate (%)",
                        "D. Minority Denial Rate", y_cap = 60)

(pp1 + pp2 + pp3 + pp4) +
  plot_layout(ncol = 4) +
  plot_annotation(
    title   = "Persistently Low vs High SIR Counties: Population and Lending Profiles (2023)",
    caption = paste0(
      "Only counties classified as persistently low (SIR < 0.75 in both 2018 and 2023) ",
      "or persistently high (SIR > 1.25 in both years) are shown. ",
      "Left facet = based on unadjusted SIR; right = risk-adjusted SIR. ",
      "Dots show individual county values overlaid on box plots."),
    theme = theme(
      plot.title   = element_text(hjust = 0.5, face = "bold", size = 11),
      plot.caption = element_text(size = 7, hjust = 0)
    )
  )
```
