---
title: "TX"
author: "Victoria"
date: "2024-10-28"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd('/Users/victoria/Library/CloudStorage/OneDrive-BowlingGreenStateUniversity/Subs/mortage_research/Analysis/mortgage-app-rates-bayesian-analysis/AssociationAnalysis')

## Package names
packages <- c('data.table',"conflicted","car","forcats","readr","tidyverse","tinytex", "datasets","papaja" ,"here","data.table",
              "DBI","RSQLite","maps","usmap","stringr","dplyr","dbplyr","knitr","forcats","parallel","psych","crunch")

## Install packages not yet installed
installed_packages <- packages %in% rownames(installed.packages())
if (any(installed_packages == FALSE)) {
  install.packages(packages[!installed_packages])
}

##Packages loading
invisible(lapply(packages, library, character.only = TRUE))
conflicts_prefer(dplyr::filter())
conflicts_prefer(crunch::combine)
conflicts_prefer(car::recode)

data_18<- fread('state_TX_2018.csv')
data_19<- fread('state_TX_2019.csv')
data_20<- fread('state_TX_2020.csv')
data_21<- fread('state_TX_2021.csv')
data_22<- fread('state_TX_2022.csv')
data_23<- fread('state_TX_2023.csv')


# Combine all the data tables into one
merged_data <- rbindlist(list(data_18, data_19, data_20, data_21, data_22, data_23), use.names = TRUE, fill = TRUE)

varnums<- function(x) {
  w=as.data.frame(c(1:length(colnames(x))),paste0('# ',colnames(x)))
  names(w)= c("# Var/Pos")
  w
}

```


# Redefine Columns

```{r data}
remove_cols<- function(df){
  w = df %>%
    filter(loan_term == "360" &
           lien_status == 1 & 
           construction_method== 1 & 
           business_or_commercial_purpose==2&
           total_units == 1 &
           loan_purpose == 1&
           occupancy_type == 1&
           conforming_loan_limit == "C"&
           derived_race != "Free Form Text Only" )
  w
}

df<- remove_cols(merged_data)

df <- df[,-c(7,9,17,18,19,21,24:38,40,41:44,45,51:59,63:66,68:73,76,77,80:83,85:88,92)]

df[,c(1:15,21:35)]<- lapply(df[,c(1:15,21:35)], factor)


conflicts_prefer(dplyr::recode)
df <- df%>%
  filter(!(action_taken %in% c("7", "8")))
df$derived_action_taken<- recode(df$action_taken, 'c("1","6") = "Approved"; c("2", "3")= "Denied"; else = "Withdrawn/Incomplete"')

df$derived_applicant_sex<- recode(df$applicant_sex,'c("1") = "Male"; c("2") = "Female";c("4") = "Not Applicable";else = "Other"')

df$derived_co_applicant_sex<- recode(df$`co-applicant_sex`,'c("1") = "Male"; c("2") = "Female";c("4") = "Not Applicable";c("5") = "No_Coapplicant"; else = "Other"')

df$denial_reason_1<- recode(df$`denial_reason-1`,'c("1") = "DTI ratio"; c("2") = "Emplmnt hist";c("3") = "Credit Hist";c("4") = "Collateral";c("5") = "Insuf Cash";c("6")="Unverifieable";c("7")="Incomplete Appl";c("8")="Insurance Denied";c("1111")="Exempt";c("10")="NA"; else = "Other"')

desired_order_age_applicant<- c('<25','25-34','35-44','45-54','55-64','65-74','>74','8888')
df$applicant_age<- factor(df$applicant_age,levels = desired_order_age_applicant)

#race

df <- df %>%
  mutate(race = ifelse(derived_ethnicity == "Joint" | derived_race == "Joint" | derived_sex == "Joint" | derived_co_applicant_sex %in% c(1,2,6),"Joint",
                       ifelse(derived_ethnicity =="Hispanic or Latino" & (`applicant_race-1` %in% c(5,6,7) | derived_race %in% c("White", "Race Not Available")), "Hispanic White",
                              ifelse(`applicant_race-1` %in% c(2,27) | derived_race == "Asian", "Asian",
                                     ifelse(derived_ethnicity =="Not Hispanic or Latino" & (`applicant_race-1` %in%c(5,6,7) | derived_race == "White") | derived_race == "White", "Non-Hispanic White",
                                            ifelse((`applicant_race-1` %in% c(1,4,21:26, 41:43) | derived_race %in% c("2 or more minority races", "American Indian or Alaska Native", "Native Hawaiian or Other Pacific Islander")), "Other Minority Races",
                                                          ifelse((`applicant_race-1` == 3 | derived_race== "Black or African American"), "Black or African American",'Missing')))))))


```


# Select features to begin with

Here, I choose to retain variables used in the initial model.

```{r}
df_2<- df%>%
  dplyr::select(-c(2:4,12:13,22:27,31,32,34,35))
```

# Removed NA values in income, DTI, LTV
I jut removed every row with missing data

```{r}
### remove income = NA
df_3<- df_2 %>%
  filter(income != 'NA')
df_3<- df_2 %>%
  filter(income != 0)
df_3<- na.omit(df_3)

df_3[, c(12:14, 21:27)] <- as.data.frame(lapply(df_3[, c(12:14, 21:27)], as.numeric))
  
```



# more variables

```{r}

df_3$income_bins<- cut(df_3$income, breaks = c(0,50, 100, 150, 200, 300, 400,500,600,1000,2000,Inf), include.lowest = TRUE)
df_3$log_income<- log(df_3$income) 

df_3$ltv_bins<- cut(df_3$loan_to_value_ratio, breaks = c(0,60,80,90,100,150,Inf), include.lowest = TRUE)

df_3$dti_bins<- ifelse(df_3$debt_to_income_ratio %in% c("<20%",'20%-<30%','30%-<36%'), '<36%',
                       ifelse(df_3$debt_to_income_ratio %in% c('36','37','38','39','40','41','42'),'36%-42%',
                              ifelse(df_3$debt_to_income_ratio %in% c('43','44','45','46','47','48','49'),'43%-49%',
                                 ifelse(df_3$debt_to_income_ratio  == '50%-60%', '50%-60%', '>60%'))))
desired_order_dti<- c('<36%','36%-42%','43%-49%','50%-60%', '>60%')
df_3$dti_bins<- factor(df_3$dti_bins,levels = desired_order_dti)
df_3$dti_g50<- ifelse(df_3$debt_to_income_ratio %in% c("<20%",'20%-<30%','30%-<36%','36','37','38','39','40','41','42',
                                                       '43','44','45','46','47','48','49'),'<50%','>=50%')

df_3$property_value_thousands<- df_3$property_value/1000
df_3$property_value_bins<- cut(df_3$property_value_thousands, breaks = c(0,200,400,500,1000,1500,Inf), include.lowest = TRUE)

df_3$loan_amt_thousands<- df_3$loan_amount/1000

## change reference category
df_3 <- within(df_3, applicant_age <- relevel(applicant_age, ref = '35-44'))
df_3 <- within(df_3, ltv_bins <- relevel(ltv_bins, ref = '(90,100]'))
df_3<- within(df_3, derived_race <- relevel(derived_race, ref = 'White'))
df_3<- within(df_3, property_value_bins <- relevel(property_value_bins, ref = '(200,400]'))
df_3<- within(df_3, derived_sex <- relevel(derived_sex, ref = 'Joint'))
df_3<- df_3[complete.cases(df_3),]


```



# list od models to loop through

```{r}

model_list <- list()

# Loop through each year to create a separate model
for (yr in unique(df_3$activity_year)) {
  yearly_data <- df_3[activity_year == yr]
  initial_model <- glm(derived_action_taken ~ 1, data = yearly_data, family = binomial)
  full_model <- glm(derived_action_taken ~ loan_amt_thousands + ffiec_msa_md_median_family_income +income_bins + race + derived_sex + tract_minority_population_percent+applicant_age +ltv_bins+ dti_g50 + property_value_bins , data = yearly_data, family = binomial)
  
  # Use stepwise forward selection; ensure doesn't print every model
  selected_model <- step(initial_model, scope = list(lower = initial_model, upper = full_model), 
                         direction = "forward",trace = 0)
  
  # Save the selected model for this year
  model_list[[as.character(yr)]] <- selected_model
}

# View summary of each model to see selected variables and coefficients
#for (yr in names(model_list)) {
#  cat("Summary for year:", yr, "\n")
#  print(summary(model_list[[yr]]))
#}


```




# results of the model

```{r}
# Set significance level threshold
significance_level <- 0.05
results <- data.frame(Year = character(),
                      Variable = character(),
                      Estimate = numeric(),
                      StdError = numeric(),
                      stringsAsFactors = FALSE)
for (yr in names(model_list)) {
  model_summary <- summary(model_list[[yr]])
  coeffs <- model_summary$coefficients
  significant_vars <- coeffs[coeffs[, "Pr(>|z|)"] < significance_level, ]
  if (nrow(significant_vars) > 0) {
    year_results <- data.frame(Year = yr,
                               Variable = rownames(significant_vars),
                               Estimate = significant_vars[, "Estimate"],
                               StdError = significant_vars[, "Std. Error"])
    
    results <- rbind(results, year_results)
  }
}

#write.csv(results, "significant_variables.csv", row.names = FALSE)

```




# Variables used in modeling

ltv
dti
income bins
property_value
race
ffiec income
sex

```{r}

# Filter data into separate data frames for each category

#using old results
results<- fread('significant_variables.csv')
  
  
income_data <- results %>%
  filter(grepl("income_bins", Variable))
race_data <- results %>%
  filter(grepl("race", Variable))
ltv_data <- results %>%
  filter(grepl("ltv_bins", Variable))
dti_data <- results %>%
  filter(grepl("dti", Variable))
property_value_data <- results %>%
  filter(grepl("property_value", Variable))
sex_data <- results %>%
  filter(grepl("sex", Variable))
loan_amt_data <- results %>%
  filter(grepl("loan", Variable))

# 1. Income Trend Plot
ggplot(income_data, aes(x = as.numeric(Year), y = Estimate, color = Variable)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  labs(title = "Trend Analysis of Log Odds of Denial by Income Over Years",
       x = "Year",
       y = "Log Odds of Denial (Estimate)",
       color = "Income Bins") +
  theme_minimal() +
  theme(legend.position = "right")

# 2. Race Trend Plot
ggplot(race_data, aes(x = as.numeric(Year), y = Estimate, color = Variable)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  labs(title = "Trend Analysis of Log Odds of Denial by Race Over Years",
       x = "Year",
       y = "Log Odds of Denial (Estimate)",
       color = "Race") +
  theme_minimal() +
  theme(legend.position = "right")

# 3. Loan-to-Value (LTV) Trend Plot
ggplot(ltv_data, aes(x = as.numeric(Year), y = Estimate, color = Variable)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  labs(title = "Trend Analysis of Log Odds of Denial by Loan-to-Value Ratio Over Years",
       x = "Year",
       y = "Log Odds of Denial (Estimate)",
       color = "LTV Bins") +
  theme_minimal() +
  theme(legend.position = "right")



# 4. DTI Trend Plot
ggplot(dti_data, aes(x = as.numeric(Year), y = Estimate, color = Variable)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  labs(title = "Trend Analysis of Log Odds of Denial by Debt-to-Income Ratio Over Years",
       x = "Year",
       y = "Log Odds of Denial (Estimate)",
       color = "DTI") +
  theme_minimal() +
  theme(legend.position = "right")

# 5. Property Value Trend Plot
ggplot(property_value_data, aes(x = as.numeric(Year), y = Estimate, color = Variable)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  labs(title = "Trend Analysis of Log Odds of Denial by Property Value Over Years",
       x = "Year",
       y = "Log Odds of Denial (Estimate)",
       color = "Property Value Bins") +
  theme_minimal() +
  theme(legend.position = "right")


# 6. Sex Denial Trend Plot
ggplot(sex_data, aes(x = as.numeric(Year), y = Estimate, color = Variable)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  labs(title = "Trend Analysis of Log Odds of Denial by Gender Over Years",
       x = "Year",
       y = "Log Odds of Denial (Estimate)",
       color = "Gender") +
  theme_minimal() +
  theme(legend.position = "right")


# 7.Loan amount Trend Plot
ggplot(loan_amt_data, aes(x = as.numeric(Year), y = Estimate, color = Variable)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  labs(title = "Trend Analysis of Log Odds of Denial by Loan Amount Over Years",
       x = "Year",
       y = "Log Odds of Denial (Estimate)",
       color = "Loan Amount") +
  theme_minimal() +
  theme(legend.position = "right")

```







# Stay Safe and rename

### May remove

```{r}

#df<- copy(training)

# df <-  df%>% 
#   rename(
#     bad_flag = 'bad_flag',
#     fico_score = 'fico',
#     amt_financed = 'amtfinanced_1req',
#     new_used = 'collateral_dlrinput_newused_1req',
#     pti = 'pti_1req',
#     ltv = 'ltv_1req',
#     gender = 'Gender',
#     race = 'Race',
#     avg_lifespan_bankcard = 'p12_bca8370_a',
#     avg_trade_lifespan = 'p12_all8370_a',
#     recent_bankcard_months = 'p12_bcc8120_a',
#     open_bankcard_5k = 'p12_bcc3456_a',
#     recent_personal_trade_months = 'p12_pil8120_a',
#     max_bal_credit_ratio = 'p12_reh7120_a',
#     max_single_bal_credit_ratio_6 = 'p12_all7120_a',
#     overall_bal_credit_ratio_bcc = 'p12_bcx7110_a',
#     auto_loan_bal_credit_ratio = 'p12_aut7110_a',
#     open_retail_bal_credit_ratio_6 = 'p12_rtr7110_a',
#     past_due_bal_ratio = 'p12_all7170_a',
#     worst_status_first_year = 'p12_all6250_a',
#     worst_trade_status = 'p12_alm6200_a',
#     worst_auto_trade = 'p12_aua6200_a',
#     never_delinquent_pct_12 = 'p12_all7937_a',
#     open_installment_pct_6 = 'p12_iln7410_a',
#     revolving_never_delinq = 'p12_rev1300_a',
#     total_trades_never_delinq = 'p12_als1300_a',
#     total_inquiries_12 = 'clact9429_a',
#     inquiries_past_90 = 'clact9428_a',
#     non_trad_inquiries_12 = 'clntr9437_a',
#     oldest_installment_months = 'p12_iln8220_a',
#     months_open_oldest_auto = 'p12_aua8220_a',
#     total_balance_trades = 'clall5010_a',
#     total_auto_trades = 'p12_aua0300_a',
#     recent_over_90_delinq_accs = 'clall2434_a',
#     recent_auto_delinquency_days = 'p12_aua8151_a',
#     trades_ever_delinq_12 = 'p12_all2427_a',
#     worst_trade_status_first_year = 'p12_all6971_a',
#     
#   )



```












