---
title: "TX – Corrected Analysis (aligned with manuscript)"
author: "Victoria"
date: "2024-10-28"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd('/Users/victoria/Library/CloudStorage/OneDrive-BowlingGreenStateUniversity/BGSU/PhDing/MortgageAnalysis')

## Package names
## FIX 1: Added 'glmnet' to package list (required for LASSO as described in manuscript Section 3.2)
packages <- c('data.table', "conflicted", "car", "forcats", "readr", "tidyverse",
              "tinytex", "datasets", "papaja", "here", "DBI", "RSQLite", "maps",
              "usmap", "stringr", "dplyr", "dbplyr", "knitr", "kableExtra",
              "forcats", "parallel", "psych", "crunch", "glmnet")

## Install packages not yet installed
installed_packages <- packages %in% rownames(installed.packages())
if (any(installed_packages == FALSE)) {
  install.packages(packages[!installed_packages])
}

## Packages loading
invisible(lapply(packages, library, character.only = TRUE))
conflicts_prefer(dplyr::filter())
conflicts_prefer(crunch::combine)
conflicts_prefer(car::recode)

data_18 <- fread('state_TX_2018.csv')
data_19 <- fread('state_TX_2019.csv')
data_20 <- fread('state_TX_2020.csv')
data_21 <- fread('state_TX_2021.csv')
data_22 <- fread('state_TX_2022.csv')
data_23 <- fread('state_TX_2023.csv')

# Combine all the data tables into one
merged_data <- rbindlist(list(data_18, data_19, data_20, data_21, data_22, data_23),
                         use.names = TRUE, fill = TRUE)

varnums <- function(x) {
  w <- as.data.frame(c(1:length(colnames(x))), paste0('# ', colnames(x)))
  names(w) <- c("# Var/Pos")
  w
}
```


# Redefine Columns

```{r data}
remove_cols <- function(df) {
  w <- df %>%
    filter(loan_term == "360" &
           lien_status == 1 &
           construction_method == 1 &
           business_or_commercial_purpose == 2 &
           total_units == 1 &
           loan_purpose == 1 &
           occupancy_type == 1 &
           conforming_loan_limit == "C" &
           derived_race != "Free Form Text Only")
  w
}

df <- remove_cols(merged_data)

df <- df[, -c(7, 9, 17, 18, 19, 21, 24:38, 40, 41:44, 45, 51:59, 63:66, 68:73, 76, 77, 80:83, 85:88, 92)]

df[, c(1:15, 21:35)] <- lapply(df[, c(1:15, 21:35)], factor)

## FIX 2: Filter out action_taken values 7 and 8 as before
df <- df %>%
  filter(!(action_taken %in% c("7", "8")))

df$derived_action_taken <- recode(df$action_taken,
  'c("1","6") = "Approved"; c("2", "3") = "Denied"; else = "Withdrawn/Incomplete"')

## FIX 2 (continued): Manuscript Section 3.1 states "Only completed applications —
## those neither withdrawn nor closed for incompleteness — are retained."
## The original code did NOT filter out Withdrawn/Incomplete before modelling.
## This step removes those rows so that the model uses only Approved vs. Denied.
df <- df %>%
  filter(derived_action_taken %in% c("Approved", "Denied"))

df$derived_applicant_sex <- recode(df$applicant_sex,
  'c("1") = "Male"; c("2") = "Female"; c("4") = "Not Applicable"; else = "Other"')

df$derived_co_applicant_sex <- recode(df$`co-applicant_sex`,
  'c("1") = "Male"; c("2") = "Female"; c("4") = "Not Applicable";
   c("5") = "No_Coapplicant"; else = "Other"')

df$denial_reason_1 <- recode(df$`denial_reason-1`,
  'c("1") = "DTI ratio"; c("2") = "Emplmnt hist"; c("3") = "Credit Hist";
   c("4") = "Collateral"; c("5") = "Insuf Cash"; c("6") = "Unverifieable";
   c("7") = "Incomplete Appl"; c("8") = "Insurance Denied";
   c("1111") = "Exempt"; c("10") = "NA"; else = "Other"')

desired_order_age_applicant <- c('<25', '25-34', '35-44', '45-54', '55-64', '65-74', '>74', '8888')
df$applicant_age <- factor(df$applicant_age, levels = desired_order_age_applicant)

# Race coding
df <- df %>%
  mutate(race = ifelse(
    derived_ethnicity == "Joint" | derived_race == "Joint" |
      derived_sex == "Joint" | derived_co_applicant_sex %in% c(1, 2, 6),
    "Joint",
    ifelse(derived_ethnicity == "Hispanic or Latino" &
             (`applicant_race-1` %in% c(5, 6, 7) |
              derived_race %in% c("White", "Race Not Available")),
           "Hispanic White",
           ifelse(`applicant_race-1` %in% c(2, 27) | derived_race == "Asian",
                  "Asian",
                  ifelse(derived_ethnicity == "Not Hispanic or Latino" &
                           (`applicant_race-1` %in% c(5, 6, 7) |
                            derived_race == "White") | derived_race == "White",
                         "Non-Hispanic White",
                         ifelse((`applicant_race-1` %in% c(1, 4, 21:26, 41:43) |
                                   derived_race %in% c("2 or more minority races",
                                                        "American Indian or Alaska Native",
                                                        "Native Hawaiian or Other Pacific Islander")),
                                "Other Minority Races",
                                ifelse(`applicant_race-1` == 3 |
                                         derived_race == "Black or African American",
                                       "Black or African American", "Missing")))))))
```


# Table 1 — Application Profile by Demography

Reproduces manuscript Table 1. Uses `df` (completed applications, before NA removal)
so application counts match the per-year totals reported in the manuscript.

```{r table1}
race_order_t1 <- c("Asian", "Black or African American", "Hispanic White",
                   "Joint", "Missing", "Non-Hispanic White", "Other Minority Races")

table1 <- df_3 %>%
  mutate(race = factor(race, levels = race_order_t1)) %>%
  group_by(activity_year, race) %>%
  summarise(
    Applications = n(),
    Denials      = sum(derived_action_taken == "Denied"),
    .groups      = "drop"
  ) %>%
  group_by(activity_year) %>%
  mutate(
    `I (%)`           = paste0(round(Applications / sum(Applications) * 100, 2), "%"),
    `III (Denial Rate)` = paste0(round(Denials / Applications * 100, 2), "%")
  ) %>%
  ungroup() %>%
  rename(
    `Year`         = activity_year,
    `Race`         = race,
    `II (Denials)` = Denials
  ) %>%
  dplyr::select(Year, Race, Applications, `I (%)`, `II (Denials)`, `III (Denial Rate)`) %>%
  arrange(Year, Race)

kable(table1,
      caption  = "Table 1: Application Profile by Demography",
      booktabs = TRUE,
      longtable = TRUE,
      align    = c("l","l","r","r","r","r")) %>%
  kable_styling(latex_options = c("repeat_header", "hold_position"),
                font_size = 9) %>%
  collapse_rows(columns = 1, valign = "top")
```


# Select features to begin with

Here, I choose to retain variables used in the initial model.

```{r}
df_2 <- df %>%
  dplyr::select(-c(2:4, 12:13, 22:27, 31, 32, 34, 35))
```

# Remove NA values in income, DTI, LTV

```{r}
df_3 <- df_2 %>%
  filter(income != 'NA')
df_3 <- df_2 %>%
  filter(income != 0)
df_3 <- na.omit(df_3)

df_3[, c(12:14, 21:27)] <- as.data.frame(lapply(df_3[, c(12:14, 21:27)], as.numeric))
```


# More variables

```{r}
df_3$income_bins <- cut(df_3$income,
  breaks = c(0, 50, 100, 150, 200, 300, 400, 500, 600, 1000, 2000, Inf),
  include.lowest = TRUE)
df_3$log_income <- log(df_3$income)

df_3$ltv_bins <- cut(df_3$loan_to_value_ratio,
  breaks = c(0, 60, 80, 90, 100, 150, Inf),
  include.lowest = TRUE)

df_3$dti_bins <- ifelse(
  df_3$debt_to_income_ratio %in% c("<20%", '20%-<30%', '30%-<36%'), '<36%',
  ifelse(df_3$debt_to_income_ratio %in% c('36','37','38','39','40','41','42'), '36%-42%',
         ifelse(df_3$debt_to_income_ratio %in% c('43','44','45','46','47','48','49'), '43%-49%',
                ifelse(df_3$debt_to_income_ratio == '50%-60%', '50%-60%', '>60%'))))
desired_order_dti <- c('<36%', '36%-42%', '43%-49%', '50%-60%', '>60%')
df_3$dti_bins <- factor(df_3$dti_bins, levels = desired_order_dti)

df_3$dti_g50 <- ifelse(
  df_3$debt_to_income_ratio %in% c("<20%", '20%-<30%', '30%-<36%',
                                    '36','37','38','39','40','41','42',
                                    '43','44','45','46','47','48','49'),
  '<50%', '>=50%')

df_3$property_value_thousands <- df_3$property_value / 1000
df_3$property_value_bins <- cut(df_3$property_value_thousands,
  breaks = c(0, 200, 400, 500, 1000, 1500, Inf),
  include.lowest = TRUE)

df_3$loan_amt_thousands <- df_3$loan_amount / 1000

## Reference categories (as specified in manuscript Section 3.2)
df_3 <- within(df_3, applicant_age       <- relevel(applicant_age,       ref = '35-44'))
df_3 <- within(df_3, ltv_bins            <- relevel(ltv_bins,            ref = '(90,100]'))
df_3 <- within(df_3, property_value_bins <- relevel(property_value_bins, ref = '(200,400]'))
df_3 <- within(df_3, derived_sex         <- relevel(derived_sex,         ref = 'Joint'))

## income_bins: default first level is [0,50]; manuscript reference is (200,300]
df_3 <- within(df_3, income_bins <- relevel(income_bins, ref = '(200,300]'))

## dti_g50: was a character vector — must be factored before releveling.
## Manuscript reference is applicants with DTI < 50%.
df_3$dti_g50 <- factor(df_3$dti_g50)
df_3 <- within(df_3, dti_g50 <- relevel(dti_g50, ref = '<50%'))

## FIX 3: The original code releveled 'derived_race' (the original HMDA column)
## but the model formula uses the custom 'race' variable. Those are different columns.
## Corrected to relevel the custom 'race' variable with reference "Non-Hispanic White",
## matching manuscript Section 3.2: "Non-Hispanic White applicants (Race)".
df_3$race <- factor(df_3$race)
df_3 <- within(df_3, race <- relevel(race, ref = 'Non-Hispanic White'))

df_3 <- df_3[complete.cases(df_3), ]

## FIX 4: Create a proper binary integer outcome variable (1 = Denied, 0 = Approved)
## as described in manuscript Section 3.2. The original code fed a factor with
## 3 levels to binomial GLM; this binary column matches the manuscript specification.
df_3$denied <- as.integer(df_3$derived_action_taken == "Denied")
```



```{r}
df_2018 <- df_3 %>% filter(activity_year == 2018)
df_2023 <- df_3 %>% filter(activity_year == 2023)

# denied
denied_18 <- sum(df_2018$denied)/nrow(df_2018)
denied_23 <- sum(df_2023$denied)/nrow(df_2023)


#
```

# Table 2 — Summary Statistics of Numerical Variables

Reproduces manuscript Table 2. Uses `df_3` (after NA removal), which contains
1,999,145 observations as reported in the manuscript.

```{r table2}
num_vars <- c('ffiec_msa_md_median_family_income',
              'loan_amt_thousands',
              'log_income',
              'property_value_thousands',
              'tract_median_age_of_housing_units',
              'tract_minority_population_percent',
              'tract_one_to_four_family_homes',
              'tract_owner_occupied_units')

var_labels <- c('Median Family Income',
                'Loan Amount (k)',
                'Log Income',
                'Property Value (k)',
                'Median Housing Age (Yrs)',
                'Minority Pop. (%)',
                '1-to-4 Family Homes',
                'Owner-Occupied Units')

table2 <- do.call(rbind, lapply(seq_along(num_vars), function(i) {
  x <- as.numeric(df_3[[num_vars[i]]])
  data.frame(
    Variable = var_labels[i],
    Max      = round(max(x,    na.rm = TRUE), 2),
    Mean     = round(mean(x,   na.rm = TRUE), 2),
    Median   = round(median(x, na.rm = TRUE), 2),
    Min      = round(min(x,    na.rm = TRUE), 2),
    S.D      = round(sd(x,     na.rm = TRUE), 2),
    stringsAsFactors = FALSE
  )
}))

kable(table2,
      caption  = "Table 2: Summary Statistics of Numerical Variables",
      booktabs = TRUE,
      align    = c("l","r","r","r","r","r"),
      row.names = FALSE) %>%
  kable_styling(latex_options = c("striped", "hold_position"), font_size = 10) %>%
  footnote(general = paste0("'k' denotes /000 for Loan Amount and Property Value. ",
                            "Minority Pop. expressed as a percentage. ",
                            "N = ", format(nrow(df_3), big.mark = ","), " observations."),
           general_title = "Note: ", footnote_as_chunk = TRUE)
```


# LASSO penalized binomial regression (per manuscript Section 3.2)

The manuscript describes penalized binomial logistic regression with LASSO
regularization using the glmnet package (Hastie et al., 2024).

**Original code issue:** The original RMD used a standard stepwise forward-selection
GLM (`glm()` + `step()`), which does not match the LASSO method described in the
manuscript. This section replaces that approach with glmnet LASSO.

```{r lasso_models}
model_list <- list()

# Predictor variables (same set as original formula)
predictor_vars <- c('loan_amt_thousands', 'ffiec_msa_md_median_family_income',
                    'income_bins', 'race', 'derived_sex',
                    'tract_minority_population_percent',
                    'applicant_age', 'ltv_bins', 'dti_g50', 'property_value_bins')

# Loop through each year
for (yr in sort(unique(df_3$activity_year))) {
  yearly_data <- df_3[df_3$activity_year == yr, ]

  # Build model matrix (glmnet requires a numeric matrix)
  x_matrix <- model.matrix(
    as.formula(paste("denied ~", paste(predictor_vars, collapse = " + "))),
    data = yearly_data
  )[, -1]  # drop intercept column

  y_vector <- yearly_data$denied

  # Cross-validated LASSO to select optimal lambda
  set.seed(123)
  cv_fit <- cv.glmnet(x = x_matrix, y = y_vector,
                      family = "binomial", alpha = 1,
                      type.measure = "mse", nfolds = 10)

  model_list[[as.character(yr)]] <- cv_fit
}



```


# Extract significant / non-zero coefficients

```{r results}
significance_level <- 0.05

results <- data.frame(
  Year     = character(),
  Variable = character(),
  Estimate = numeric(),
  stringsAsFactors = FALSE
)

for (yr in names(model_list)) {
  cv_fit <- model_list[[yr]]

  # Use lambda.min (smallest CV error); lambda.1se gives sparser model
  coefs <- coef(cv_fit, s = "lambda.min")
  coef_df <- as.data.frame(as.matrix(coefs))
  colnames(coef_df) <- "Estimate"
  coef_df$Variable <- rownames(coef_df)
  # Drop intercept, zero-shrunk coefficients, and any NA estimates
  # (NA != 0 returns NA which passes the filter, so !is.na() guard is needed)
  coef_df <- coef_df[coef_df$Variable != "(Intercept)" &
                     !is.na(coef_df$Estimate) &
                     coef_df$Estimate != 0, ]

  if (nrow(coef_df) > 0) {
    year_results <- data.frame(
      Year     = yr,
      Variable = coef_df$Variable,
      Estimate = coef_df$Estimate
    )
    results <- rbind(results, year_results)
  }
}

# write.csv(results, "significant_variables_lasso.csv", row.names = FALSE)
```


# Post-LASSO GLM — Significance Testing

LASSO selects variables but does not produce p-values. Following standard
post-selection inference practice, we refit a standard `glm()` per year using
only the variables LASSO retained (non-zero at lambda.min). This recovers
standard errors, z-values, and p-values, allowing significance stars to be
reported as in manuscript Table 3.

```{r post_lasso_glm}

# Significance stars helper
sig_stars <- function(p) {
  case_when(
    p < 0.001 ~ "***",
    p < 0.01  ~ "**",
    p < 0.05  ~ "*",
    p < 0.1   ~ ".",
    TRUE       ~ ""
  )
}

glm_results <- list()   # full glm objects, one per year
sig_results <- data.frame(
  Year      = character(),
  Variable  = character(),
  Estimate  = numeric(),
  StdError  = numeric(),
  Pvalue    = numeric(),
  Stars     = character(),
  stringsAsFactors = FALSE
)

for (yr in names(model_list)) {
  cv_fit      <- model_list[[yr]]
  yearly_data <- df_3[df_3$activity_year == yr, ]

  # Step 1: identify variables LASSO kept (non-zero at lambda.min)
  lasso_coefs <- coef(cv_fit, s = "lambda.min")
  kept_vars   <- rownames(lasso_coefs)[
    !is.na(lasso_coefs[, 1]) &
    lasso_coefs[, 1] != 0 &
    rownames(lasso_coefs) != "(Intercept)"
  ]

  if (length(kept_vars) == 0) next

  # Step 2: map dummy names back to original predictor variable names.
  # model.matrix expands factors into dummies like "dti_g50>=50%" or
  # "raceBlack or African American". We recover the base variable name by
  # checking which predictor_vars each dummy name starts with.
  # A regex like ([a-z_]+) breaks on variables with digits (e.g. dti_g50).
  base_vars <- predictor_vars[sapply(predictor_vars, function(v) {
    any(startsWith(kept_vars, v))
  })]

  # Build formula from kept base variables
  glm_formula <- as.formula(
    paste("denied ~", paste(base_vars, collapse = " + "))
  )

  # Step 3: refit standard GLM with selected variables
  glm_fit <- glm(glm_formula, data = yearly_data, family = binomial)
  glm_results[[yr]] <- glm_fit

  # Step 4: extract coefficients, SE, and p-values
  coef_summary <- summary(glm_fit)$coefficients
  year_sig <- data.frame(
    Year     = yr,
    Variable = rownames(coef_summary),
    Estimate = round(coef_summary[, "Estimate"],    3),
    StdError = round(coef_summary[, "Std. Error"],  3),
    Pvalue   = round(coef_summary[, "Pr(>|z|)"],    4),
    Stars    = sig_stars(coef_summary[, "Pr(>|z|)"]),
    stringsAsFactors = FALSE
  )
  sig_results <- rbind(sig_results, year_sig)
}
```


# Table 3 — LASSO Logistic Regression Coefficients (2018–2023)

Reproduces manuscript Table 3 in wide format (one column per year). LASSO
shrinks coefficients toward zero rather than producing p-values; non-zero
coefficients are retained at lambda.min. Cells show the LASSO log-odds
estimate; zeroed-out coefficients are shown as "—".

```{r table3}

# Clean up variable names to match manuscript row labels

clean_varname <- function(x) {
  x <- gsub("income_bins",        "Income(K)",        x)
  x <- gsub("race",               "Race",             x)
  x <- gsub("derived_sex",        "Sex",              x)
  x <- gsub("ltv_bins",           "LTV(%)",           x)
  x <- gsub("applicant_age",      "Age",              x)
  x <- gsub("dti_g50",            "DTI(%)",           x)
  x <- gsub("property_value_bins","Property Value(K)", x)
  x <- gsub("loan_amt_thousands", "Loan Amount(K)",   x)
  x <- gsub("ffiec_msa_md_median_family_income", "MSA Income", x)
  x <- gsub("tract_minority_population_percent", "Minority Pop.(%)", x)
  x
}

table3_long <- sig_results %>%
  filter(Variable != "(Intercept)") %>%
  mutate(
    Variable    = clean_varname(Variable),
    # Combine estimate and stars into one cell, e.g. "1.216***"
    Est_Stars   = paste0(Estimate, Stars)
  )

table3_wide <- table3_long %>%
  dplyr::select(Year, Variable, Est_Stars) %>%
  pivot_wider(names_from = Year, values_from = Est_Stars) %>%
  arrange(Variable) %>%
  mutate(across(-Variable, ~ifelse(is.na(.), "—", .)))

# Add intercept row from post-LASSO GLM (includes p-value and stars)
intercepts <- sig_results %>%
  filter(Variable == "(Intercept)") %>%
  mutate(Est_Stars = paste0(Estimate, Stars)) %>%
  dplyr::select(Year, Variable, Est_Stars)

intercept_wide <- intercepts %>%
  pivot_wider(names_from = Year, values_from = Est_Stars) %>%
  mutate(across(-Variable, ~ifelse(is.na(.), "—", .)))

table3_final <- bind_rows(intercept_wide, table3_wide)

kable(table3_final,
      caption   = "Table 3: LASSO Logistic Regression Coefficients (2018–2023)",
      booktabs  = TRUE,
      align     = c("l", rep("r", 6)),
      row.names = FALSE) %>%
  kable_styling(latex_options = c("striped", "hold_position", "scale_down"),
                font_size = 8) %>%
  footnote(general = paste0("Coefficients are log-odds estimates from a binomial GLM refitted on variables ",
                            "selected by 10-fold cross-validated LASSO (lambda.min). ",
                            "A dash (—) indicates the variable was shrunk to zero by LASSO and excluded from the model for that year. ",
                            "Significance levels: *** p<0.001; ** p<0.01; * p<0.05; . p<0.1. ",
                            "Reference categories: Non-Hispanic White (Race), Joint (Sex), ",
                            "35–44 (Age), LTV 90–100%, DTI <50%, Income $200k–$300k, Property Value $200k–$400k."),
           general_title = "Note: ", footnote_as_chunk = TRUE)
```


# Model accuracy and MSE (per manuscript Section 3.3)

```{r model_fit}
# NOTE: mean(pred_prob) on the *training* data always equals mean(y_obs) because
# glmnet does not penalize the intercept — the intercept's optimality condition
# forces sum(fitted) = sum(observed). To obtain a meaningful predicted denial rate
# that can differ from the unadjusted rate, we evaluate on a held-out test set.

set.seed(123)
fit_stats <- data.frame(
  Year               = character(),
  Accuracy           = numeric(),
  MSE                = numeric(),
  Predicted_Rate_pct = numeric(),   # mean predicted denial prob on TEST set × 100
  Observed_Rate_pct  = numeric(),   # actual denial rate on TEST set × 100
  stringsAsFactors   = FALSE
)

model_list_train <- list()   # retrained models on 80 % split (for out-of-sample eval)

for (yr in names(model_list)) {
  yearly_data <- df_3[df_3$activity_year == yr, ]

  # 80 / 20 train-test split (same seed → reproducible)
  n        <- nrow(yearly_data)
  train_ix <- sample(seq_len(n), size = floor(0.8 * n))
  train    <- yearly_data[ train_ix, ]
  test     <- yearly_data[-train_ix, ]

  # Build matrices
  make_x <- function(dat) {
    model.matrix(
      as.formula(paste("denied ~", paste(predictor_vars, collapse = " + "))),
      data = dat
    )[, -1]
  }

  x_train <- make_x(train);  y_train <- train$denied
  x_test  <- make_x(test);   y_test  <- test$denied

  # Refit LASSO on training fold
  cv_fit_train <- cv.glmnet(x = x_train, y = y_train,
                             family = "binomial", alpha = 1,
                             type.measure = "mse", nfolds = 10)
  model_list_train[[yr]] <- cv_fit_train

  # Predict on held-out TEST set
  pred_prob  <- as.vector(predict(cv_fit_train, newx = x_test,
                                  s = "lambda.min", type = "response"))
  pred_class <- as.integer(pred_prob >= 0.5)

  accuracy  <- mean(pred_class == y_test)
  mse       <- mean((y_test - pred_prob)^2)
  pred_rate <- mean(pred_prob)   # out-of-sample: can differ from observed rate
  obs_rate  <- mean(y_test)

  fit_stats <- rbind(fit_stats,
                     data.frame(
                       Year               = yr,
                       Accuracy           = round(accuracy  * 100, 2),
                       MSE                = round(mse,             4),
                       Predicted_Rate_pct = round(pred_rate * 100, 2),
                       Observed_Rate_pct  = round(obs_rate  * 100, 2)
                     ))
}

# Print full table
print(fit_stats)

# Quick inline checks for 2018 and 2023
for (yr in c("2018", "2023")) {
  row <- fit_stats[fit_stats$Year == yr, ]
  if (nrow(row) > 0)
    cat(sprintf(
      "%s — Accuracy: %.2f%%  |  MSE: %.4f  |  Predicted rate: %.2f%%  |  Observed rate: %.2f%%\n",
      yr, row$Accuracy, row$MSE, row$Predicted_Rate_pct, row$Observed_Rate_pct))
}


```


# Trend plots

```{r trend_plots}
# Use results from above (or load saved CSV)
# results <- fread('significant_variables_lasso.csv')


income_data         <- results %>% filter(grepl("income_bins",     Variable))
race_data           <- results %>% filter(grepl("race",            Variable))
ltv_data            <- results %>% filter(grepl("ltv_bins",        Variable))
dti_data            <- results %>% filter(grepl("dti",             Variable))
property_value_data <- results %>% filter(grepl("property_value",  Variable))
sex_data            <- results %>% filter(grepl("sex",             Variable))
loan_amt_data       <- results %>% filter(grepl("loan",            Variable))

# 1. Income Trend Plot
ggplot(income_data, aes(x = as.numeric(Year), y = Estimate, color = Variable)) +
  geom_line(size = 1) + geom_point(size = 2) +
  labs(title = "Trend Analysis of Log Odds of Denial by Income Over Years",
       x = "Year", y = "Log Odds of Denial (Estimate)", color = "Income Bins") +
  theme_minimal() + theme(legend.position = "right")

# 2. Race Trend Plot
ggplot(race_data, aes(x = as.numeric(Year), y = Estimate, color = Variable)) +
  geom_line(size = 1) + geom_point(size = 2) +
  labs(title = "Trend Analysis of Log Odds of Denial by Race Over Years",
       x = "Year", y = "Log Odds of Denial (Estimate)", color = "Race") +
  theme_minimal() + theme(legend.position = "right")

# 3. LTV Trend Plot
ggplot(ltv_data, aes(x = as.numeric(Year), y = Estimate, color = Variable)) +
  geom_line(size = 1) + geom_point(size = 2) +
  labs(title = "Trend Analysis of Log Odds of Denial by Loan-to-Value Ratio Over Years",
       x = "Year", y = "Log Odds of Denial (Estimate)", color = "LTV Bins") +
  theme_minimal() + theme(legend.position = "right")

# 4. DTI Trend Plot
ggplot(dti_data, aes(x = as.numeric(Year), y = Estimate, color = Variable)) +
  geom_line(size = 1) + geom_point(size = 2) +
  labs(title = "Trend Analysis of Log Odds of Denial by Debt-to-Income Ratio Over Years",
       x = "Year", y = "Log Odds of Denial (Estimate)", color = "DTI") +
  theme_minimal() + theme(legend.position = "right")

# 5. Property Value Trend Plot
ggplot(property_value_data, aes(x = as.numeric(Year), y = Estimate, color = Variable)) +
  geom_line(size = 1) + geom_point(size = 2) +
  labs(title = "Trend Analysis of Log Odds of Denial by Property Value Over Years",
       x = "Year", y = "Log Odds of Denial (Estimate)", color = "Property Value Bins") +
  theme_minimal() + theme(legend.position = "right")

# 6. Sex Denial Trend Plot
ggplot(sex_data, aes(x = as.numeric(Year), y = Estimate, color = Variable)) +
  geom_line(size = 1) + geom_point(size = 2) +
  labs(title = "Trend Analysis of Log Odds of Denial by Gender Over Years",
       x = "Year", y = "Log Odds of Denial (Estimate)", color = "Gender") +
  theme_minimal() + theme(legend.position = "right")

# 7. Loan Amount Trend Plot
ggplot(loan_amt_data, aes(x = as.numeric(Year), y = Estimate, color = Variable)) +
  geom_line(size = 1) + geom_point(size = 2) +
  labs(title = "Trend Analysis of Log Odds of Denial by Loan Amount Over Years",
       x = "Year", y = "Log Odds of Denial (Estimate)", color = "Loan Amount") +
  theme_minimal() + theme(legend.position = "right")
```


---

# Appendix A2 — Denial Rate by Year

```{r appendix_a2}

app_a2 <- df_3 %>%
  mutate(activity_year = as.integer(as.character(activity_year))) %>%
  group_by(activity_year) %>%
  summarise(
    `Number of Applications` = n(),
    `Denial Rate`            = paste0(
      round(mean(derived_action_taken == "Denied") * 100, 2), "%"),
    .groups = "drop"
  ) %>%
  arrange(activity_year) %>%
  mutate(`#` = row_number()) %>%
  dplyr::select(`#`, Year = activity_year,
                `Number of Applications`, `Denial Rate`)

kable(app_a2,
      caption   = "Appendix A2: Denial Rate by Year",
      booktabs  = TRUE,
      align     = c("r","r","r","r"),
      row.names = FALSE) %>%
  kable_styling(latex_options = c("striped", "hold_position"), font_size = 10)

```
